---
format:
  live-html:
    toc: true
---
# Regression Case Study: NAPLAN Reading Scores {#sec-regression-case-study-naplan-reading-scores}

## Defining the research question {#sec-defining-the-research-question}

- Investigate how student and school-level [predictors]{.glossary term="Predictor (explanatory) variable"} relate to NAPLAN
  Reading scores. Clarify [outcome]{.glossary term="Response (outcome) variable"}, units, and any grouping structure.

## Importing data {#sec-importing-data}

```{r module05-case-naplan-importing-data-load-data}
#| message: false
# Replace path with the appropriate CSV or RDS from the Resources/naplan reading folder
# naplan <- read.csv("Resources/naplan reading/naplan_reading.csv")
```

## Exploratory plot {#sec-exploratory-plot}

- Start with scatterplots and boxplots for key predictors (e.g., study
  time, SES, gender) against Reading scores.

```{r module05-case-naplan-exploratory-plot-setup}
# Example placeholder using mtcars structure; swap to naplan data
library(ggplot2)
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(x = "Predictor", y = "Reading score")
```

## Fitting the [least-squares]{.glossary term="Ordinary least squares"} model {#sec-fitting-the-least-squares-model}

- Begin with a first-order additive model; consider [interactions]{.glossary term="Interaction"} or
  [polynomial terms]{.glossary term="Polynomial term"} if exploratory plots suggest curvature.

```{r module05-case-naplan-fitting-the-least-squares-model-fit-lm}
# model <- lm(Reading ~ predictor1 + predictor2, data = naplan)
# summary(model)
```

## Model refinement and interpretation {#sec-model-refinement-and-interpretation}

- Check residual diagnostics (see Module 6). Address nonlinearity,
  [heteroskedasticity]{.glossary}, or [influential points]{.glossary term="Influential point"}.
- Interpret coefficients, including categorical contrasts and any
  interaction terms.

```{r module05-case-naplan-model-refinement-and-interpretation-summary}
# plot(model, which = 1:2)
# coef(summary(model))
```

## Reporting the results {#sec-reporting-the-results}

- Present fitted effects with [confidence intervals]{.glossary term="Confidence interval"}, and provide a
  [prediction interval]{.glossary term="Prediction interval"} for a meaningful scenario (e.g., a student profile
  of interest).
- Summarise key findings in plain language and note any limitations
  (e.g., omitted variables, sample size).
