---
format:
  live-html:
    toc: true
execute:
  echo: false
  warning: false
  message: false

embed-resources: true

filters:
  - custom-numbered-blocks

custom-numbered-blocks:
  classes:
    Assumption:
      colors: ["#FFCDD2", "#F44336"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      boxstyle: foldbox.simple
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false

webr:
  render-df: gt-interactive

resources:
  - "Data/naplan_reading.csv"
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}



```{r ch-case-naplan-setup-r}
#| include: false
library(MASS)
library(tidyverse)
library(broom)
theme_set(theme_bw())

naplan <- read.csv("Data/naplan_reading.csv") |>
  as_tibble() |>
  mutate(
    grade = factor(grade, levels = c("Year 3", "Year 5", "Year 7", "Year 9")),
    parent_education = factor(
      parent_education,
      levels = c(
        "Year 10 or below", "Year 12", "Certificate/Diploma",
        "Bachelor degree", "Postgraduate"
      )
    ),
    school_type = factor(school_type, levels = c("Government", "Catholic", "Independent")),
    gender = factor(gender, levels = c("Female", "Male")),
    birth_months = factor(
      birth_months,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    )
  )

naplan_n <- nrow(naplan)
```

```{r ch-case-naplan-models}
#| include: false
#---- SLR and stepwise MLR build-up (for §5 and §6)
mod_slr <- lm(naplan_reading_score ~ ses_index, data = naplan)
mod_ses_grade <- lm(naplan_reading_score ~ ses_index + grade, data = naplan)
mod_cats <- lm(
  naplan_reading_score ~ ses_index + grade + school_type + gender + parent_education,
  data = naplan
)

#---- First-order "everything we have" (as a screening start)
mod_full_first_order <- lm(
  naplan_reading_score ~ grade + ses_index + reading_time_home + n_siblings +
    parent_education + school_type + gender + birth_months,
  data = naplan
)

#---- Candidate refinements guided by EDA + model-building tools (see @sec-a-model-building-workflow)
mod_no_birth <- update(mod_full_first_order, . ~ . - birth_months)
mod_quad_time <- update(mod_no_birth, . ~ . - reading_time_home + poly(reading_time_home, 2, raw = TRUE))
mod_quad_time_int_grade <- lm(
  naplan_reading_score ~ grade * poly(reading_time_home, 2, raw = TRUE) +
    ses_index + n_siblings + parent_education + school_type + gender,
  data = naplan
)

#---- Two ways to represent “SES” in the model (see discussion later)
mod_ses_index <- lm(
  naplan_reading_score ~ grade * poly(reading_time_home, 2, raw = TRUE) +
    school_type + gender + ses_index,
  data = naplan
)
mod_parent_edu <- lm(
  naplan_reading_score ~ grade * poly(reading_time_home, 2, raw = TRUE) +
    school_type + gender + parent_education,
  data = naplan
)
mod_parent_edu_plus_ses <- update(mod_parent_edu, . ~ . + ses_index)

#---- Final choice for this chapter: keep parent education as the SES proxy
final_mod <- mod_parent_edu
final_glance <- glance(final_mod)
final_fstat <- summary(final_mod)$fstatistic
final_p_overall <- pf(final_fstat[1], final_fstat[2], final_fstat[3], lower.tail = FALSE)

fit_tbl <- tibble(
  model = c(
    "Full first-order (+ birth month)",
    "Drop birth month",
    "Add quadratic reading time",
    "Add grade × reading time (quadratic)",
    "Add SES index to parent education",
    "Final (use parent education; drop SES index)"
  ),
  object = list(
    mod_full_first_order, mod_no_birth, mod_quad_time, mod_quad_time_int_grade,
    mod_parent_edu_plus_ses, final_mod
  )
) |>
  mutate(
    formula = map_chr(object, \(m) paste(deparse(formula(m)), collapse = " ")),
    glance = map(object, glance)
  ) |>
  unnest(glance) |>
  transmute(
    model,
    formula,
    nobs,
    adj_r2 = adj.r.squared,
    rse = sigma,
    AIC,
    BIC
  )

final_terms <- tidy(final_mod, conf.int = TRUE)
```

# Regression Case Study: NAPLAN Reading Scores {#sec-regression-case-study-naplan-reading-scores}

In @sec-introduction-to-linear-models we built the basic linear model, in @sec-slr and @sec-multiple-linear-regression-mlr we focused on fitting and interpretation, in @sec-model-building-and-selection we built and compared candidate models, and in @sec-regression-pitfalls-and-diagnostics we diagnosed problems. In this final chapter, we put those pieces together in an end-to-end analysis.

Our workflow follows the model-building approach in @sec-a-model-building-workflow:

1. Define the research question and candidate variables.
2. Explore the data (plots + summaries).
3. Fit a reasonable starting model.
4. Refine using model comparisons (without “blind selection”).
5. Check diagnostics and interpret results.
6. Write up the findings in plain language.

This chapter is designed as a **hands-on walkthrough**. The webR chunks are the analysis: run each one as you read. Only use “Skip exercise” if you’ve already completed that step (e.g. in your IDE) and just want to keep moving.

## Defining the research question {#sec-defining-the-research-question}

This case study uses `naplan_reading.csv` (in this project: `Data/naplan_reading.csv`), which contains NAPLAN reading achievement data for **3,000 Australian students** across **Years 3, 5, 7, and 9**, drawn from **60 schools** (`school_id`). Each row is one student.

The outcome variable is `naplan_reading_score`. Because scores are on different ranges across grades (roughly **100–600** in Year 3 vs **400–900** in Year 9), we include `grade` as a covariate throughout.

Alongside the outcome, the dataset includes student and family predictors such as weekly reading time at home (in minutes), parent education, school type, gender, a socioeconomic status index, birth month, and number of siblings.

**Research question:** How are students’ socioeconomic circumstances, reading practice at home, and demographic characteristics associated with NAPLAN Reading scores?

In this dataset:

- The [outcome]{.glossary term="Outcome (response) variable"} is `naplan_reading_score`.
- Candidate [predictors]{.glossary term="Predictor (explanatory) variable"} include:
  - reading practice: `reading_time_home` (minutes per week),
  - socioeconomic measures: `ses_index` and `parent_education`,
  - demographics: `gender`, `birth_months`, `n_siblings`,
  - schooling context: `school_type`,
  - schooling level: `grade` (Year 3/5/7/9, included as a covariate).

::: Note
This is an observational dataset. Regression estimates here describe **associations**, not “effects”. For reminders about modelling assumptions, see @sec-common-assumptions-and-pitfalls.
:::

::: Note
Students are clustered within schools (`school_id`). If there are unmeasured school-level factors (e.g. teaching practices, resources, peer effects), then students from the same school can end up with **correlated errors**, which violates the “independent errors” assumption used for standard t/F tests in a plain `lm()`.

In this chapter we stay within ordinary least squares (no random effects). We *partly* address school-to-school differences by including observed context variables such as `school_type`, along with strong student-level covariates like `grade` and socioeconomic proxies. These predictors can explain some between-school variation and reduce residual clustering, but they do not guarantee independence.

A more complete approach would use either:

- `lm(... + factor(school_id))` (school fixed effects): controls for all stable between-school differences, but uses many parameters and you generally can’t estimate purely school-level predictors (like `school_type`) at the same time, or
- a multilevel model with a random effect for `school_id`, or
- cluster-robust standard errors.

These options are beyond the scope of this short course, but are worth knowing about if you analyse clustered education data.
:::

## Importing and preparing the data {#sec-importing-data}

We start by importing the dataset and setting factor levels so coefficient interpretations are clear (see @sec-mlr_cat and @sec-reference-levels-and-releveling).

```{webr ch-case-naplan-setup}
#| include: false
library(tidyverse)
library(broom)
theme_set(theme_bw())
```
```{webr ch-case-naplan-models-webr}
#| include: false
#| output: false
#| edit: false
naplan <- read_csv("Data/naplan_reading.csv", show_col_types = FALSE) |>
  mutate(
    grade = factor(grade, levels = c("Year 3", "Year 5", "Year 7", "Year 9")),
    parent_education = factor(
      parent_education,
      levels = c(
        "Year 10 or below",
        "Year 12",
        "Certificate/Diploma",
        "Bachelor degree",
        "Postgraduate"
      )
    ),
    school_type = factor(school_type, levels = c("Government", "Catholic", "Independent")),
    gender = factor(gender, levels = c("Female", "Male")),
    birth_months = factor(
      birth_months,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    )
  )

naplan_n <- nrow(naplan)

#---- SLR and stepwise MLR build-up (for §5 and §6)
mod_slr <- lm(naplan_reading_score ~ ses_index, data = naplan)
mod_ses_grade <- lm(naplan_reading_score ~ ses_index + grade, data = naplan)
mod_cats <- lm(
  naplan_reading_score ~ ses_index + grade + school_type + gender + parent_education,
  data = naplan
)

#---- First-order "everything we have" (as a screening start)
mod_full_first_order <- lm(
  naplan_reading_score ~ grade + ses_index + reading_time_home + n_siblings +
    parent_education + school_type + gender + birth_months,
  data = naplan
)

#---- Candidate refinements guided by EDA + model-building tools (see @sec-a-model-building-workflow)
mod_no_birth <- update(mod_full_first_order, . ~ . - birth_months)
mod_quad_time <- update(mod_no_birth, . ~ . - reading_time_home + poly(reading_time_home, 2, raw = TRUE))
mod_quad_time_int_grade <- lm(
  naplan_reading_score ~ grade * poly(reading_time_home, 2, raw = TRUE) +
    ses_index + n_siblings + parent_education + school_type + gender,
  data = naplan
)

#---- Two ways to represent “SES” in the model (see discussion later)
mod_ses_index <- lm(
  naplan_reading_score ~ grade * poly(reading_time_home, 2, raw = TRUE) +
    school_type + gender + ses_index,
  data = naplan
)
mod_parent_edu <- lm(
  naplan_reading_score ~ grade * poly(reading_time_home, 2, raw = TRUE) +
    school_type + gender + parent_education,
  data = naplan
)
mod_parent_edu_plus_ses <- update(mod_parent_edu, . ~ . + ses_index)

#---- Final choice for this chapter: keep parent education as the SES proxy
final_mod <- mod_parent_edu
final_aug <- augment(final_mod)

fit_tbl <- tibble(
  model = c(
    "Full first-order (+ birth month)",
    "Drop birth month",
    "Add quadratic reading time",
    "Add grade × reading time (quadratic)",
    "Add SES index to parent education",
    "Final (use parent education; drop SES index)"
  ),
  object = list(
    mod_full_first_order, mod_no_birth, mod_quad_time, mod_quad_time_int_grade,
    mod_parent_edu_plus_ses, final_mod
  )
) |>
  mutate(
    formula = map_chr(object, function(m) paste(deparse(formula(m)), collapse = " ")),
    glance = map(object, glance)
  ) |>
  unnest(glance) |>
  transmute(
    model,
    formula,
    nobs,
    adj_r2 = adj.r.squared,
    rse = sigma,
    AIC,
    BIC
  )

final_terms <- tidy(final_mod, conf.int = TRUE)
```

::: Exercise
Run the chunk below to import and prepare the data. The `glimpse()` output shows all available variables, and the small dictionary table summarises what each variable represents.

When you’re done, click **Check** to continue. If you’ve already done this step (e.g. in your IDE), click **Skip exercise** instead.

::: {.cell exercise='ex_5.1' envir='Ex5'}
```{webr}
#| exercise: ex_5.1
#| envir: Ex5
#| autorun: true
naplan <- read_csv("Data/naplan_reading.csv", show_col_types = FALSE) |>
  mutate(
    grade = factor(grade, levels = c("Year 3", "Year 5", "Year 7", "Year 9")),
    parent_education = factor(
      parent_education,
      levels = c(
        "Year 10 or below",
        "Year 12",
        "Certificate/Diploma",
        "Bachelor degree",
        "Postgraduate"
      )
    ),
    school_type = factor(school_type, levels = c("Government", "Catholic", "Independent")),
    gender = factor(gender, levels = c("Female", "Male")),
    birth_months = factor(
      birth_months,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    )
  )

glimpse(naplan)

naplan_dictionary <- tribble(
  ~variable, ~description,
  "student_id", "Student identifier",
  "school_id", "School identifier (60 schools)",
  "grade", "Year level (Year 3/5/7/9)",
  "reading_time_home", "Minutes of reading at home per week",
  "parent_education", "Parents' highest education level",
  "school_type", "School sector (Government/Catholic/Independent)",
  "gender", "Student gender",
  "birth_months", "Birth month (Jan–Dec)",
  "n_siblings", "Number of siblings",
  "ses_index", "Socioeconomic status index (continuous)",
  "naplan_reading_score", "NAPLAN reading score (outcome)"
)

naplan_dictionary

naplan |>
  select(
    student_id, school_id, grade, naplan_reading_score, ses_index, reading_time_home,
    parent_education, school_type, gender, birth_months, n_siblings
  ) |>
  slice_head(n = 8)

naplan |>
  summarise(
    n = n(),
    min_score = min(naplan_reading_score),
    max_score = max(naplan_reading_score),
    mean_score = mean(naplan_reading_score),
    sd_score = sd(naplan_reading_score),
    n_schools = n_distinct(school_id)
  )

naplan |>
  count(grade) |>
  arrange(grade)
```
:::

:::: {.solution exercise="ex_5.1"}
::: {.cell exercise='ex_5.1' solution='true'}
```{webr}
#| exercise: ex_5.1
#| solution: true
naplan <- read_csv("Data/naplan_reading.csv", show_col_types = FALSE) |>
  mutate(
    grade = factor(grade, levels = c("Year 3", "Year 5", "Year 7", "Year 9")),
    parent_education = factor(
      parent_education,
      levels = c(
        "Year 10 or below",
        "Year 12",
        "Certificate/Diploma",
        "Bachelor degree",
        "Postgraduate"
      )
    ),
    school_type = factor(school_type, levels = c("Government", "Catholic", "Independent")),
    gender = factor(gender, levels = c("Female", "Male")),
    birth_months = factor(
      birth_months,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    )
  )

glimpse(naplan)

naplan_dictionary <- tribble(
  ~variable, ~description,
  "student_id", "Student identifier",
  "school_id", "School identifier (60 schools)",
  "grade", "Year level (Year 3/5/7/9)",
  "reading_time_home", "Minutes of reading at home per week",
  "parent_education", "Parents' highest education level",
  "school_type", "School sector (Government/Catholic/Independent)",
  "gender", "Student gender",
  "birth_months", "Birth month (Jan–Dec)",
  "n_siblings", "Number of siblings",
  "ses_index", "Socioeconomic status index (continuous)",
  "naplan_reading_score", "NAPLAN reading score (outcome)"
)

naplan_dictionary

naplan |>
  select(
    student_id, school_id, grade, naplan_reading_score, ses_index, reading_time_home,
    parent_education, school_type, gender, birth_months, n_siblings
  ) |>
  slice_head(n = 8)

naplan |>
  summarise(
    n = n(),
    min_score = min(naplan_reading_score),
    max_score = max(naplan_reading_score),
    mean_score = mean(naplan_reading_score),
    sd_score = sd(naplan_reading_score),
    n_schools = n_distinct(school_id)
  )

naplan |>
  count(grade) |>
  arrange(grade)
```
:::
::::

::: {.cell exercise='ex_5.1' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_5.1
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  if (!exists("naplan", envir = .envir_result)) {
    fail("Run the chunk above so an object named `naplan` exists.")
  }
  d <- get("naplan", envir = .envir_result)
  if (!inherits(d, "data.frame")) fail("`naplan` should be a data frame/tibble.")
  if (nrow(d) != 3000) fail("Expected 3000 rows (one per student).")

  required <- c(
    "student_id", "school_id", "grade", "reading_time_home",
    "parent_education", "school_type", "gender", "birth_months",
    "n_siblings", "ses_index", "naplan_reading_score"
  )
  missing <- setdiff(required, names(d))
  if (length(missing) > 0) {
    fail(paste0("Missing columns: ", paste(missing, collapse = ", ")))
  }

  if (!identical(levels(d$grade), c("Year 3", "Year 5", "Year 7", "Year 9"))) {
    fail("Set `grade` levels to: Year 3, Year 5, Year 7, Year 9.")
  }
  if (!identical(
    levels(d$parent_education),
    c("Year 10 or below", "Year 12", "Certificate/Diploma", "Bachelor degree", "Postgraduate")
  )) {
    fail("Set `parent_education` levels to match the data dictionary (Year 10 or below → Postgraduate).")
  }
  if (!identical(levels(d$school_type), c("Government", "Catholic", "Independent"))) {
    fail("Set `school_type` levels to: Government, Catholic, Independent.")
  }
  if (!identical(levels(d$gender), c("Female", "Male"))) {
    fail("Set `gender` levels to: Female, Male.")
  }
  if (!identical(levels(d$birth_months), c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))) {
    fail("Set `birth_months` levels to Jan–Dec.")
  }

  pass("Nice — data imported and cleaned. You’re ready for EDA.")
})
```
:::
:::

::: Exercise
As a first quick EDA summary, create a table called `score_by_grade` with the mean NAPLAN reading score for each grade, sorted from highest to lowest.

When it looks right, click **Check** to continue (or **Skip exercise** if you did this in your IDE).

::: {.cell exercise='ex_5.2' envir='Ex5' define='["score_by_grade"]'}
```{webr}
#| exercise: ex_5.2
#| envir: Ex5
#| autorun: true
#| define:
#|   - score_by_grade
score_by_grade <- naplan |>
  group_by(grade) |>
  summarise(mean_score = mean(naplan_reading_score)) |>
  arrange(desc(mean_score))

score_by_grade
```
:::

:::: {.solution exercise="ex_5.2"}
::: {.cell exercise='ex_5.2' solution='true'}
```{webr}
#| exercise: ex_5.2
#| solution: true
score_by_grade <- naplan |>
  group_by(grade) |>
  summarise(mean_score = mean(naplan_reading_score)) |>
  arrange(desc(mean_score))

score_by_grade
```
:::
::::

::: {.cell exercise='ex_5.2' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_5.2
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  if (!exists("score_by_grade", envir = .envir_result)) {
    fail("Create a table named `score_by_grade`.")
  }
  s <- get("score_by_grade", envir = .envir_result)
  if (!inherits(s, "data.frame")) fail("`score_by_grade` should be a data frame/tibble.")
  if (!all(c("grade", "mean_score") %in% names(s))) fail("Include columns named `grade` and `mean_score`.")
  if (nrow(s) != 4) fail("Expected one row per grade (4 rows).")
  if (!isTRUE(all.equal(s$mean_score, sort(s$mean_score, decreasing = TRUE)))) {
    fail("Make sure `score_by_grade` is sorted from highest mean score to lowest.")
  }
  if (as.character(s$grade[1]) != "Year 9") {
    fail("Hint: the top row should be Year 9 once sorted descending.")
  }
  pass("Nice — that’s a good first summary before we move to plots.")
})
```
:::
:::

## Exploratory data analysis {#sec-exploratory-plot}

Before we fit any model, we look for:

- obvious trends (linear vs curved),
- differences across groups (categorical predictors),
- potential [multicollinearity]{.glossary} (predictors that tell us almost the same thing),
- outliers or unusual values that might affect the model.

This is exactly the “EDA first” mindset in @sec-exploratory-data-analysis-and-theory-driven-model-development.

### Clustering by school (students within schools)

Because students are sampled within schools, a quick EDA step is to check the *cluster sizes* (how many students per school) and whether school-average scores differ. Visible variation across schools is one sign that school context might matter, which in turn can lead to correlated errors within schools if that context is not fully captured by the predictors. Later, when we include `school_type` (and other predictors) in the regression, we are adjusting for observed school context, which can reduce—but not eliminate—between-school clustering in the residuals.

```{webr ch-case-naplan-eda-school-sizes}
#| echo: true
#| autorun: true
naplan |>
  count(school_id, name = "n_students") |>
  summarise(
    n_schools = n(),
    min_students = min(n_students),
    median_students = median(n_students),
    max_students = max(n_students)
  )
```

```{webr ch-case-naplan-eda-school-means}
#| echo: true
#| autorun: true
school_means <- naplan |>
  group_by(school_id) |>
  summarise(
    n_students = n(),
    mean_score = mean(naplan_reading_score),
    .groups = "drop"
  )

ggplot(school_means, aes(x = mean_score)) +
  geom_histogram(binwidth = 20, fill = "steelblue", colour = "white") +
  labs(x = "School mean NAPLAN reading score", y = "Number of schools")
```

### Outcome distribution (by grade)

NAPLAN scores are on different ranges across grades (Year 3 scores are typically lower than Year 9 scores), so it helps to look grade-by-grade.

```{webr ch-case-naplan-eda-score-by-grade}
#| echo: true
#| autorun: true
ggplot(naplan, aes(x = naplan_reading_score)) +
  geom_histogram(binwidth = 20, fill = "steelblue", colour = "white") +
  facet_wrap(~ grade, ncol = 2, scales = "free_y") +
  labs(x = "NAPLAN reading score", y = "Number of students")
```

### Key bivariate relationships

#### Reading score vs SES index

```{webr ch-case-naplan-eda-ses}
#| echo: true
#| autorun: true
ggplot(naplan, aes(x = ses_index, y = naplan_reading_score, colour = grade)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "SES index", y = "NAPLAN reading score", colour = "Grade")
```

#### Reading score vs time spent reading at home

Because `reading_time_home` is reported in minutes per week and has a lot of small values (including zeros), we should be open to nonlinearity here (see @sec-non_linearity).

```{webr ch-case-naplan-eda-reading-time}
#| echo: true
#| autorun: true
ggplot(naplan, aes(x = reading_time_home, y = naplan_reading_score, colour = grade)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2, raw = TRUE), se = FALSE) +
  labs(x = "Reading time at home (minutes/week)", y = "NAPLAN reading score", colour = "Grade")
```

### Categorical predictors

Boxplots make it easy to compare distributions across categories (see @sec-mlr_cat and @sec-multi-level-factors-and-anova).

```{webr ch-case-naplan-eda-cats}
#| echo: true
#| autorun: true
ggplot(naplan, aes(x = parent_education, y = naplan_reading_score)) +
  geom_boxplot(outlier.alpha = 0.15) +
  coord_flip() +
  labs(x = "Parent education (highest)", y = "NAPLAN reading score")

ggplot(naplan, aes(x = school_type, y = naplan_reading_score)) +
  geom_boxplot(outlier.alpha = 0.15) +
  labs(x = "School type", y = "NAPLAN reading score")

ggplot(naplan, aes(x = gender, y = naplan_reading_score)) +
  geom_boxplot(outlier.alpha = 0.15) +
  labs(x = "Gender", y = "NAPLAN reading score")
```

### Correlations among continuous variables

We don’t have `GGally::ggpairs()` available in this project, but we can still compute a correlation matrix and look at a basic pairs plot (see @sec-pairs-plots-and-correlation-matrices).

```{webr ch-case-naplan-eda-cor}
#| echo: true
#| autorun: true
naplan_cont <- naplan |>
  select(naplan_reading_score, ses_index, reading_time_home, n_siblings) |>
  mutate(across(everything(), as.numeric))

round(cor(naplan_cont), 2)
```

```{webr ch-case-naplan-eda-pairs}
#| echo: true
#| autorun: true
pairs(
  naplan_cont,
  pch = 19,
  cex = 0.35,
  col = scales::alpha("black", 0.25)
)
```

::: Exercise
Recreate one of the key EDA plots: reading time at home vs NAPLAN reading score, with a quadratic smoother and separate panels for each grade.

After it runs, try experimenting (e.g., set `se = FALSE`, remove `facet_wrap(~ grade)`, or change the smoother to linear).

::: {.cell exercise='ex_5.3' envir='Ex5' define='["p_readtime"]'}
```{webr}
#| exercise: ex_5.3
#| envir: Ex5
#| autorun: true
#| define:
#|   - p_readtime
p_readtime <- ggplot(naplan, aes(x = reading_time_home, y = naplan_reading_score)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2, raw = TRUE), se = TRUE) +
  facet_wrap(~ grade) +
  labs(x = "Reading time at home (minutes/week)", y = "NAPLAN reading score")

p_readtime
```
:::

:::: {.solution exercise="ex_5.3"}
::: {.cell exercise='ex_5.3' solution='true'}
```{webr}
#| exercise: ex_5.3
#| solution: true
p_readtime <- ggplot(naplan, aes(x = reading_time_home, y = naplan_reading_score)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2, raw = TRUE), se = TRUE) +
  facet_wrap(~ grade) +
  labs(x = "Reading time at home (minutes/week)", y = "NAPLAN reading score")

p_readtime
```
:::
::::

::: {.cell exercise='ex_5.3' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_5.3
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  p <- get("p_readtime", envir = .envir_result)
  if (!inherits(p, "ggplot")) fail("Create a ggplot object named `p_readtime`.")
  tryCatch(ggplot_build(p), error = function(e) {
    fail("Your plot doesn't build yet. Hint: check brackets/commas and that `naplan` is spelled correctly.")
  })
  if (inherits(p$facet, "FacetNull")) fail("Add a facet, e.g. `facet_wrap(~ grade)`.")
  pass("Nice — you’ve built a faceted plot showing the reading-time relationship by grade.")
})
```
:::
:::

## Starting simple: a single-predictor model {#sec-case-naplan-slr}

Before jumping to a complex model, let's start with the simplest possible regression — a single predictor. This revisits the core SLR skills from @sec-slr: fitting a line, interpreting the slope, quantifying uncertainty, and recognising limitations.

From the EDA, `ses_index` has the highest correlation with score among the continuous predictors ($r \approx 0.21$). Let's see how far that one predictor gets us.

### Fitting the SLR model

```{webr ch-case-naplan-slr-fit}
#| echo: true
#| autorun: true
mod_slr <- lm(naplan_reading_score ~ ses_index, data = naplan)
summary(mod_slr)
```

```{webr ch-case-naplan-slr-tidy}
#| echo: true
#| autorun: true
tidy(mod_slr, conf.int = TRUE)
glance(mod_slr)
```

Let's interpret these results (see @sec-estimating-parameters and @sec-beta_inference):

- **Intercept** ($\approx 547$): a student with `ses_index = 0` (roughly average SES) is predicted to score about 547 points.
- **Slope** ($\approx 22$): each 1-unit increase in the SES index is associated with about 22 additional points on the NAPLAN reading test.
- **95% CI for slope** ($\approx [18, 25]$): we're confident the SES–score association is positive.
- **$t = 11.8$, $p < 0.001$**: strong evidence that the association is not zero.

But look at the fit statistics:

- $R^2 \approx 0.045$: SES alone explains only about 4.5% of the variation in scores.
- RSE $\approx 92$: typical prediction errors are about 92 points — very large relative to within-grade standard deviations (~60–70 points).

::: Key-point
Statistical significance ($p < 0.001$) does not mean practical sufficiency. A highly significant predictor can still leave most of the variation unexplained. We clearly need more predictors.
:::

### Prediction with a single predictor

What does a [prediction interval]{.glossary term="Prediction interval"} look like with this simple model? Consider a student with above-average SES (`ses_index = 1.5`):

```{webr ch-case-naplan-slr-predict}
#| echo: true
#| autorun: true
new_student <- tibble(ses_index = 1.5)

# Confidence interval for the MEAN response at ses_index = 1.5
predict(mod_slr, newdata = new_student, interval = "confidence")

# Prediction interval for a NEW student at ses_index = 1.5
predict(mod_slr, newdata = new_student, interval = "prediction")
```

The prediction interval is enormous (~±180 points) because the RSE is so large. This is the situation described in @sec-inference-about-the-response-y: when $R^2$ is low, individual predictions are very uncertain even when the slope is statistically significant.

### Why one predictor isn't enough

```{webr ch-case-naplan-slr-why-more}
#| echo: true
#| autorun: true
ggplot(naplan, aes(x = ses_index, y = naplan_reading_score, colour = grade)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "SES index", y = "NAPLAN reading score", colour = "Grade")
```

The four roughly parallel lines tell us that `grade` explains a large portion of the remaining variation. Students in higher grades score higher regardless of SES. This is exactly the situation where [multiple linear regression]{.glossary term="Multiple linear regression"} helps — we can adjust for grade and get a cleaner picture of the SES association *within* each grade level.

::: Exercise
Fit a simple linear regression of `naplan_reading_score ~ reading_time_home` and store it as `mod_slr_rt`. Then look at the slope and $R^2$.

The slope tells you how much the predicted score increases per additional minute of weekly reading. Is it large? Is $R^2$ any better than the SES model?

::: {.cell exercise='ex_cs_slr' envir='Ex5' define='["mod_slr_rt"]'}
```{webr}
#| exercise: ex_cs_slr
#| envir: Ex5
#| autorun: true
#| define:
#|   - mod_slr_rt
mod_slr_rt <- lm(naplan_reading_score ~ reading_time_home, data = naplan)
tidy(mod_slr_rt, conf.int = TRUE)
glance(mod_slr_rt) |> select(r.squared, sigma, AIC)
```
:::

:::: {.solution exercise="ex_cs_slr"}
::: {.cell exercise='ex_cs_slr' solution='true'}
```{webr}
#| exercise: ex_cs_slr
#| solution: true
mod_slr_rt <- lm(naplan_reading_score ~ reading_time_home, data = naplan)
tidy(mod_slr_rt, conf.int = TRUE)
glance(mod_slr_rt) |> select(r.squared, sigma, AIC)
```
:::
::::

::: {.cell exercise='ex_cs_slr' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_cs_slr
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  if (!exists("mod_slr_rt", envir = .envir_result)) fail("Create an object named `mod_slr_rt`.")
  m <- get("mod_slr_rt", envir = .envir_result)
  if (!inherits(m, "lm")) fail("`mod_slr_rt` should be an `lm` model.")

  resp <- as.character(formula(m)[[2]])
  if (resp != "naplan_reading_score") fail("The response variable should be `naplan_reading_score`.")

  rhs <- attr(terms(m), "term.labels")
  if (!("reading_time_home" %in% rhs) || length(rhs) != 1) {
    fail("Fit a *simple* regression with only `reading_time_home` as predictor.")
  }
  pass("Good — the slope is about 0.6 points per minute, and R² ≈ 0.02. Even weaker than SES alone! The EDA suggested curvature here, so a quadratic might help later.")
})
```
:::
:::

::: Note
The slope of ~0.6 points per extra minute of weekly reading looks small. But remember: (1) this is a *linear* fit to a possibly *curved* relationship (we saw diminishing returns in the EDA), and (2) confounders like grade aren't adjusted for yet. We'll return to this predictor with a polynomial term in @sec-model-refinement-and-interpretation.
:::

## Extending to multiple regression {#sec-case-naplan-mlr}

We've seen that single predictors leave most variation unexplained. Now let's build up a multiple regression model step-by-step, adding predictors and watching how the fit changes. This revisits the skills from @sec-multiple-linear-regression-mlr: partial coefficients, categorical predictors, reference levels, and overall model tests.

### Adding grade: the dominant predictor

Grade explains the largest share of variation because NAPLAN scores are on different scales across year levels.

```{webr ch-case-naplan-mlr-grade}
#| echo: true
#| autorun: true
mod_ses_grade <- lm(naplan_reading_score ~ ses_index + grade, data = naplan)
glance(mod_ses_grade) |> select(adj.r.squared, sigma, AIC)
tidy(mod_ses_grade, conf.int = TRUE)
```

Look at what happened:

- **Adjusted $R^2$** jumped from 0.044 to roughly 0.60 — adding `grade` explains most of the previously unexplained variation.
- **RSE** dropped from ~92 to ~60 — predictions are now much tighter.
- The **SES slope** barely changed (~22 → ~21) — `ses_index` and `grade` are nearly uncorrelated, so their effects are largely independent (see @sec-partial-regression-coefficients-betai).
- **Grade coefficients** (relative to Year 3, the reference): Year 5 ≈ +83, Year 7 ≈ +137, Year 9 ≈ +184 points. These are [indicator variable]{.glossary term="Indicator (dummy) variable"} coefficients — exactly the dummy coding from @sec-factors-and-indicator-variables.

### Adding more categorical predictors

Let's bring in `school_type`, `gender`, and `parent_education`:

```{webr ch-case-naplan-mlr-cats}
#| echo: true
#| autorun: true
mod_cats <- lm(
  naplan_reading_score ~ ses_index + grade + school_type + gender + parent_education,
  data = naplan
)
glance(mod_cats) |> select(adj.r.squared, sigma, AIC)
```

```{webr ch-case-naplan-mlr-cats-coefs}
#| echo: true
#| autorun: true
tidy(mod_cats, conf.int = TRUE) |> print(n = 20)
```

The reference levels are (see @sec-reference-levels-and-releveling):

- `grade`: Year 3
- `school_type`: Government
- `gender`: Female
- `parent_education`: Year 10 or below

So each coefficient represents the estimated difference from that baseline, holding everything else constant.

### The overall F-test

The [F-test]{.glossary term="F-test (overall model test)"} asks: "Does this model explain significantly more variation than an intercept-only model?" (see @sec-the-f-test-for-model-model-usefulness-and-measures-of-model-fit).

```{webr ch-case-naplan-mlr-ftest}
#| echo: true
#| autorun: true
mod_cats_summary <- summary(mod_cats)
cat("F-statistic:", round(mod_cats_summary$fstatistic[1], 1), 
    "on", mod_cats_summary$fstatistic[2], "and", mod_cats_summary$fstatistic[3], "df\n")
cat("p-value:", format.pval(pf(mod_cats_summary$fstatistic[1], mod_cats_summary$fstatistic[2], 
    mod_cats_summary$fstatistic[3], lower.tail = FALSE), digits = 3), "\n")
```

With this many observations and this many significant predictors, the F-test is overwhelmingly significant. The model as a whole clearly "works" — the real question is *which* predictors we need and *how* to specify them.

::: Exercise
From the `tidy()` output of `mod_cats` above, find the coefficient for `parent_educationPostgraduate`. Store it (just the numeric estimate) as `coef_postgrad`. Then answer: what does this number mean in context?

::: {.cell exercise='ex_cs_interpret_coef' envir='Ex5' define='["coef_postgrad"]'}
```{webr}
#| exercise: ex_cs_interpret_coef
#| envir: Ex5
#| autorun: true
#| define:
#|   - coef_postgrad
coef_postgrad <- tidy(mod_cats) |>
  filter(term == "parent_educationPostgraduate") |>
  pull(estimate)

coef_postgrad
```
:::

:::: {.solution exercise="ex_cs_interpret_coef"}
::: {.cell exercise='ex_cs_interpret_coef' solution='true'}
```{webr}
#| exercise: ex_cs_interpret_coef
#| solution: true
coef_postgrad <- tidy(mod_cats) |>
  filter(term == "parent_educationPostgraduate") |>
  pull(estimate)

coef_postgrad

# Interpretation: Students whose parents have a Postgraduate degree are predicted
# to score approximately `coef_postgrad` points HIGHER on the NAPLAN reading test
# than students whose parents have Year 10 or below education (the reference level),
# holding grade, SES index, school type, and gender constant.
```
:::
::::

::: {.cell exercise='ex_cs_interpret_coef' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_cs_interpret_coef
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  if (!exists("coef_postgrad", envir = .envir_result)) fail("Create an object named `coef_postgrad`.")
  ans <- get("coef_postgrad", envir = .envir_result)
  if (!is.numeric(ans) || length(ans) != 1) fail("`coef_postgrad` should be a single number.")

  expected <- coef(mod_cats)["parent_educationPostgraduate"]
  if (isTRUE(all.equal(ans, expected, tolerance = 1e-4))) {
    pass("Correct! This coefficient tells you the estimated difference in score between Postgraduate and Year 10 or below (the reference), holding other predictors constant.")
  }
  fail("Not quite. Hint: filter the tidy() output for the term 'parent_educationPostgraduate' and pull the estimate.")
})
```
:::
:::

### Do we need both ses_index and parent_education?

Both variables capture aspects of socioeconomic status. Including both may introduce [multicollinearity]{.glossary term="Multicollinearity"} — they partly tell the model the same thing (see @sec-multicollinearity-diagnostics).

```{webr ch-case-naplan-mlr-ses-overlap}
#| echo: true
#| autorun: true
mod_no_ses <- update(mod_cats, . ~ . - ses_index)
mod_no_parent <- update(mod_cats, . ~ . - parent_education)

tibble(
  model = c("Both (SES + parent_ed)", "Drop ses_index", "Drop parent_education"),
  adj_r2 = c(
    glance(mod_cats)$adj.r.squared,
    glance(mod_no_ses)$adj.r.squared,
    glance(mod_no_parent)$adj.r.squared
  ),
  AIC = c(AIC(mod_cats), AIC(mod_no_ses), AIC(mod_no_parent))
) |> print()
```

Dropping `ses_index` barely hurts — `parent_education` captures most of the SES-related signal. We'll revisit this choice formally in the model-selection section below.

## Fitting the full first-order model {#sec-fitting-the-least-squares-model}

We now fit a first-order additive model with *all* candidate predictors, including `birth_months` and `n_siblings`. This is a common starting point: fit everything, then refine (see @sec-a-model-building-workflow and @sec-increasing-model-complexity-to-model-complex-relationships).

### A first-order screening model

This model includes *all* candidate predictors, including `birth_months`. Think of it as a starting point for refinement (not the final model).

```{webr ch-case-naplan-fit-full}
#| echo: true
#| autorun: true
formula(mod_full_first_order)
```

```{webr ch-case-naplan-fit-full-glance}
#| echo: true
#| autorun: true
glance(mod_full_first_order)
```

### Model comparison: does birth month help?

We compare the model with and without `birth_months` using AIC/BIC (see @sec-information-criteria-aic-bic) and an ANOVA comparison (see @sec-comparing-model-fit).

```{webr ch-case-naplan-compare-birth-month}
#| echo: true
#| autorun: true
anova(mod_no_birth, mod_full_first_order)
fit_tbl |>
  filter(model %in% c("Full first-order (+ birth month)", "Drop birth month")) |>
  print()
```

The comparison suggests `birth_months` doesn’t meaningfully improve the model here, so we drop it.

::: Exercise
Fit the first-order “screening” model yourself, and store it as `my_full_mod`.

::: {.cell exercise='ex_5.4' envir='Ex5' define='["my_full_mod"]'}
```{webr}
#| exercise: ex_5.4
#| envir: Ex5
#| autorun: true
#| define:
#|   - my_full_mod
my_full_mod <- lm(
  naplan_reading_score ~ grade + ses_index + reading_time_home + n_siblings +
    parent_education + school_type + gender + birth_months,
  data = naplan
)

glance(my_full_mod)
```
:::

:::: {.solution exercise="ex_5.4"}
::: {.cell exercise='ex_5.4' solution='true'}
```{webr}
#| exercise: ex_5.4
#| solution: true
my_full_mod <- lm(
  naplan_reading_score ~ grade + ses_index + reading_time_home + n_siblings +
    parent_education + school_type + gender + birth_months,
  data = naplan
)

glance(my_full_mod)
```
:::
::::

::: {.cell exercise='ex_5.4' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_5.4
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  if (!exists("my_full_mod", envir = .envir_result)) fail("Create an object named `my_full_mod`.")
  m <- get("my_full_mod", envir = .envir_result)
  if (!inherits(m, "lm")) fail("`my_full_mod` should be an `lm` model.")

  rhs <- attr(terms(m), "term.labels")
  needed <- c("grade", "ses_index", "reading_time_home", "n_siblings",
              "parent_education", "school_type", "gender", "birth_months")
  if (!all(needed %in% rhs)) fail("Hint: check that you included all candidate predictors.")
  pass("Great — you’ve fitted the full first-order model.")
})
```
:::
:::

## Model refinement and selection {#sec-model-refinement-and-interpretation}

Now we refine the model in a controlled way, guided by what we saw in the EDA and by the model-building tools in @sec-model-building-and-selection.

::: Note
Automated methods like stepwise regression can be useful as **screening tools**, but can be misleading if used blindly (see @sec-stepwise-regression and @sec-blind-model-selection). In this case study we prefer small, interpretable comparisons.
:::

### (Optional) Stepwise AIC screening

If you want to see what a purely automated AIC search does, we can run a stepwise procedure on a **first-order** model (no interactions or polynomial terms). Treat this as a *starting point for thought*, not a final answer.

```{webr ch-case-naplan-stepwise}
#| echo: true
#| edit: false
#| eval: false
scope_first_order <- list(
  lower = ~ 1,
  upper = ~ grade + ses_index + reading_time_home + n_siblings +
    parent_education + school_type + gender + birth_months
)

step_mod <- step(
  lm(naplan_reading_score ~ 1, data = naplan),
  scope = scope_first_order,
  direction = "both",
  trace = 0
)

formula(step_mod)
glance(step_mod)
```

### Add a quadratic term for reading time

The scatterplot suggested possible curvature in the relationship between `reading_time_home` and `naplan_reading_score`. A natural next step is a quadratic term (see @sec-non_linearity).

```{webr ch-case-naplan-compare-quad}
#| echo: true
#| autorun: true
anova(mod_no_birth, mod_quad_time)

fit_tbl |>
  filter(model %in% c("Drop birth month", "Add quadratic reading time")) |>
  print()
```

### Allow the reading-time curve to differ by grade (interaction)

Because `grade` is strongly related to score, it’s plausible the reading-time pattern differs across grades. This is a continuous × categorical [interaction]{.glossary term="Interaction"} (see @sec-continuous-by-categorical-interactions).

```{webr ch-case-naplan-compare-interaction}
#| echo: true
#| autorun: true
anova(mod_quad_time, mod_quad_time_int_grade)

fit_tbl |>
  filter(model %in% c("Add quadratic reading time", "Add grade × reading time (quadratic)")) |>
  print()
```

### Two competing “SES” choices: SES index vs parent education

The dataset contains *two* SES-related variables:

- `ses_index` (a numeric socioeconomic index),
- `parent_education` (a categorical proxy related to SES).

Including both can be redundant: they are different ways of measuring similar underlying circumstances. This is a good moment to practise the “partial effects” idea from @sec-partial-regression-coefficients-betai, and to remember the warning about multicollinearity in @sec-multicollinearity-diagnostics.

We fit two versions:

- `mod_ses_index`: includes `ses_index` (but not `parent_education`),
- `final_mod`: includes `parent_education` (but not `ses_index`).

```{webr ch-case-naplan-ses-compare}
#| echo: true
#| autorun: true
fit_ses_tbl <- tibble(
  model = c("SES index model", "Parent education model (final)"),
  object = list(mod_ses_index, final_mod)
) |>
  mutate(glance = map(object, glance)) |>
  unnest(glance) |>
  transmute(model, adj_r2 = adj.r.squared, AIC, BIC)

fit_ses_tbl |>
  print()
```

In this dataset, the parent-education version fits better by AIC/BIC, so we use it as our final model for reporting. But we still report what the SES index suggests later, as a sensitivity check.

Does the SES index add anything once parent education is already in the model?

```{webr ch-case-naplan-ses-added-value}
#| echo: true
#| autorun: true
anova(final_mod, mod_parent_edu_plus_ses)

fit_tbl |>
  filter(model %in% c("Final (use parent education; drop SES index)",
                      "Add SES index to parent education")) |>
  print()
```

::: Exercise
Use an ANOVA comparison to test whether adding `birth_months` improves the first-order model. Store the p-value in `p_birth`.

::: {.cell exercise='ex_5.5' envir='Ex5' define='["p_birth"]'}
```{webr}
#| exercise: ex_5.5
#| envir: Ex5
#| autorun: true
#| define:
#|   - p_birth
birth_cmp <- anova(mod_no_birth, mod_full_first_order)
p_birth <- birth_cmp$`Pr(>F)`[2]
p_birth
```
:::

:::: {.solution exercise="ex_5.5"}
::: {.cell exercise='ex_5.5' solution='true'}
```{webr}
#| exercise: ex_5.5
#| solution: true
birth_cmp <- anova(mod_no_birth, mod_full_first_order)
p_birth <- birth_cmp$`Pr(>F)`[2]
p_birth
```
:::
::::

::: {.cell exercise='ex_5.5' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_5.5
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  if (!exists("p_birth", envir = .envir_result)) fail("Create an object named `p_birth`.")
  ans <- get("p_birth", envir = .envir_result)
  if (!is.numeric(ans) || length(ans) != 1) fail("`p_birth` should be a single number.")

  expected <- anova(mod_no_birth, mod_full_first_order)$`Pr(>F)`[2]
  if (isTRUE(all.equal(ans, expected, tolerance = 1e-8))) {
    pass("Nice — you’ve extracted the ANOVA p-value correctly.")
  }
  fail("Not quite. Hint: the p-value is in the second row of the ANOVA table.")
})
```
:::
:::

## Diagnostics: checking each assumption {#sec-case-naplan-diagnostics}

In @sec-regression-pitfalls-and-diagnostics we learned to check assumptions systematically using residual plots. Now we apply that framework to our final model. We check each assumption in turn: linearity, constant variance, normality, independence, and influence (see @sec-common-diagnostic-plots and @sec-diagnosing-issues-and-choosing-remedies).

```{webr ch-case-naplan-diagnostics-setup}
#| echo: true
#| autorun: true
#| edit: false
final_aug <- augment(final_mod)
final_aug$student_id <- naplan$student_id
final_aug$school_id <- naplan$school_id
```

### Linearity (residuals vs fitted values)

The residuals-vs-fitted plot is our primary tool for detecting missed curvature (see @sec-diagnosing-nonlinearity). If the model captures the systematic patterns well, residuals should scatter randomly around zero with no visible trend.

```{webr ch-case-naplan-diagnostics-linearity}
#| echo: true
#| autorun: true
ggplot(final_aug, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "gray40") +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "loess", se = FALSE, colour = "red", linewidth = 0.8) +
  labs(x = "Fitted value", y = "Residual",
       title = "Residuals vs Fitted")
```

The red smoother stays close to zero — no strong curvature remains. The quadratic term for `reading_time_home` and the `grade × reading_time` interaction appear to have captured the main nonlinearity.

We can also check residuals against the predictor that was most likely to show curvature:

```{webr ch-case-naplan-diagnostics-resid-vs-readtime}
#| echo: true
#| autorun: true
final_aug$reading_time_home <- naplan$reading_time_home

ggplot(final_aug, aes(x = reading_time_home, y = .resid)) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "loess", se = FALSE, colour = "red") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "Reading time at home (min/wk)", y = "Residual",
       title = "Residuals vs reading_time_home")
```

If this plot showed a clear curve, we'd need a higher-order polynomial or a different functional form. Here, the loess line stays near zero — the quadratic is adequate.

### Constant variance / homoscedasticity (scale-location plot)

The [scale-location plot]{.glossary term="Homoscedasticity"} checks whether the spread of residuals changes with the fitted value (see @sec-diagnosing-heteroskedasticity). We plot $\sqrt{|\text{standardised residual}|}$ against fitted values — a flat smoother indicates constant variance.

```{webr ch-case-naplan-diagnostics-scale-location}
#| echo: true
#| autorun: true
ggplot(final_aug, aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.2, size = 1) +
  geom_smooth(method = "loess", se = FALSE, colour = "red", linewidth = 0.8) +
  labs(x = "Fitted value", y = expression(sqrt("|Standardised residual|")),
       title = "Scale-Location plot")
```

If the red line trends upward (or downward), variance is changing. With $n = 3000$, mild heteroskedasticity affects standard errors but not coefficient estimates. If it were severe, we might use robust/sandwich standard errors (beyond this course).

### Normality of residuals (QQ plot + histogram)

The [QQ plot]{.glossary term="Normal Q-Q plot"} compares the distribution of standardised residuals to a theoretical normal distribution (see @sec-assessing-normality-of-residuals).

```{webr ch-case-naplan-diagnostics-qq}
#| echo: true
#| autorun: true
ggplot(final_aug, aes(sample = .std.resid)) +
  stat_qq(alpha = 0.25) +
  stat_qq_line(colour = "gray40") +
  labs(x = "Theoretical quantiles", y = "Standardised residuals",
       title = "Normal Q-Q plot")
```

```{webr ch-case-naplan-diagnostics-hist}
#| echo: true
#| autorun: true
ggplot(final_aug, aes(x = .std.resid)) +
  geom_histogram(binwidth = 0.25, fill = "steelblue", colour = "white") +
  labs(x = "Standardised residual", y = "Count",
       title = "Histogram of standardised residuals")
```

::: Key-point
With $n = 3000$, moderate departures from normality have little impact on coefficient estimates or standard errors (central limit theorem). We mainly look for extreme skew or very heavy tails — these would affect prediction intervals and small-sample inference.
:::

### Independence: residual clustering by school

Students within the same school may share unmeasured influences (teaching quality, school culture, resources). If these aren't captured by our predictors, residuals for students in the same school will be correlated — violating the independence assumption (see @sec-assessing-independence).

Let's check whether school-level mean residuals show more variation than we'd expect by chance:

```{webr ch-case-naplan-diagnostics-school-clustering}
#| echo: true
#| autorun: true
school_resid <- final_aug |>
  group_by(school_id) |>
  summarise(
    mean_resid = mean(.resid),
    n_students = n(),
    .groups = "drop"
  ) |>
  arrange(mean_resid)

ggplot(school_resid, aes(x = reorder(school_id, mean_resid), y = mean_resid)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(x = "School (sorted by mean residual)", y = "Mean residual",
       title = "Average residual by school") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{webr ch-case-naplan-diagnostics-school-sd}
#| echo: true
#| autorun: true
cat("SD of school-level mean residuals:", round(sd(school_resid$mean_resid), 2), "\n")
cat("If independent, expected SD ≈ RSE / sqrt(median school size) ≈",
    round(summary(final_mod)$sigma / sqrt(median(school_resid$n_students)), 2), "\n")
```

If the observed SD of school-mean residuals is substantially larger than the "independent" benchmark, that confirms residual clustering.

::: Note
Our model reduces clustering by including `school_type`, but cannot eliminate it. The practical consequence: our standard errors and p-values are likely *slightly too optimistic*. In a thorough analysis, we'd use cluster-robust standard errors or a mixed-effects model with a random intercept for `school_id`. These approaches are beyond this short course — but this is exactly the kind of limitation you should note when reporting results from clustered data.
:::

### Influential observations (Cook's distance)

[Cook's distance]{.glossary term="Cook's distance"} measures how much the fitted model would change if a single observation were removed (see @sec-standardised-residuals-leverage-cook-s-distance).

```{webr ch-case-naplan-diagnostics-cooksd}
#| echo: true
#| autorun: true
ggplot(final_aug, aes(x = seq_len(nrow(final_aug)), y = .cooksd)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_hline(yintercept = 4 / nrow(final_aug), linetype = "dashed", colour = "red") +
  labs(x = "Observation index", y = "Cook's distance",
       title = "Cook's distance",
       caption = paste("Threshold line: 4/n =", round(4 / nrow(final_aug), 5)))
```

```{webr ch-case-naplan-diagnostics-influence-table}
#| echo: true
#| autorun: true
final_aug |>
  arrange(desc(.cooksd)) |>
  select(student_id, school_id, grade, naplan_reading_score, .fitted, .std.resid, .hat, .cooksd) |>
  slice_head(n = 10) |>
  print()
```

With $n = 3000$, even the largest Cook's D values are small in absolute terms. No single observation dominates the regression. We'll verify this with a sensitivity analysis in the next section.

### Multicollinearity check (VIF)

Because VIFs can be inflated by raw polynomials and interactions, we compute VIFs on the simpler first-order model (no quadratic or interaction terms) to get a clearer sense of redundancy among predictors (see @sec-multicollinearity-diagnostics).

```{webr ch-case-naplan-diagnostics-vif}
#| echo: true
#| autorun: true
vif_from_lm <- function(model) {
  x <- model.matrix(model)
  x <- x[, colnames(x) != "(Intercept)", drop = FALSE]
  if (ncol(x) == 0) return(tibble(term = character(), vif = numeric()))
  if (ncol(x) == 1) return(tibble(term = colnames(x), vif = 1))

  vifs <- map_dbl(seq_len(ncol(x)), function(j) {
    r2 <- summary(lm(x[, j] ~ x[, -j]))$r.squared
    1 / (1 - r2)
  })

  tibble(term = colnames(x), vif = vifs) |>
    arrange(desc(vif))
}

vif_from_lm(mod_no_birth) |>
  slice_head(n = 12) |>
  print()
```

::: Key-point
A rule of thumb: VIF > 5–10 suggests problematic multicollinearity. If VIFs are low, predictors provide largely independent information and coefficient estimates are stable.
:::

### Diagnostics summary

| Assumption | Check | Verdict |
|-----------|-------|---------|
| Linearity | Residuals vs fitted | ✓ No strong curvature (quadratic term handles it) |
| Constant variance | Scale-location plot | Mild changes in spread — note but not severe |
| Normality | QQ plot + histogram | Approximately normal; minor tail deviations |
| Independence | School-mean residuals | Some residual clustering by school — a known limitation |
| Influence | Cook's distance | No dominant observations |
| Multicollinearity | VIF | Acceptable levels |

::: Exercise
Using `augment()` on the final model, create a residuals-vs-fitted plot. Add a loess smoother (in red) to help visualise any remaining pattern. Store the plot as `p_resid`.

::: {.cell exercise='ex_cs_diag' envir='Ex5' define='["p_resid"]'}
```{webr}
#| exercise: ex_cs_diag
#| envir: Ex5
#| autorun: true
#| define:
#|   - p_resid
p_resid <- ggplot(final_aug, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "gray40") +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "loess", se = FALSE, colour = "red") +
  labs(x = "Fitted value", y = "Residual")

p_resid
```
:::

:::: {.solution exercise="ex_cs_diag"}
::: {.cell exercise='ex_cs_diag' solution='true'}
```{webr}
#| exercise: ex_cs_diag
#| solution: true
p_resid <- ggplot(final_aug, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "gray40") +
  geom_point(alpha = 0.25, size = 1) +
  geom_smooth(method = "loess", se = FALSE, colour = "red") +
  labs(x = "Fitted value", y = "Residual")

p_resid
```
:::
::::

::: {.cell exercise='ex_cs_diag' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_cs_diag
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  p <- get("p_resid", envir = .envir_result)
  if (!inherits(p, "ggplot")) fail("Create a ggplot object named `p_resid`.")

  x <- rlang::as_label(p$mapping$x)
  y <- rlang::as_label(p$mapping$y)
  if (!(identical(x, ".fitted") && identical(y, ".resid"))) {
    fail("Hint: use `aes(x = .fitted, y = .resid)` in your `ggplot()` call.")
  }
  pass("Nice — a loess smoother near zero confirms linearity is satisfied. This is the single most important diagnostic plot (see @sec-common-diagnostic-plots).")
})
```
:::
:::

## Sensitivity analysis: do influential observations matter? {#sec-case-naplan-sensitivity}

In @sec-handling-outliers-and-influential-observations we discussed that influential observations can change model conclusions. A simple sensitivity check is to refit the model *without* the most influential points and see whether key coefficients change.

```{webr ch-case-naplan-sensitivity-refit}
#| echo: true
#| autorun: true
# Identify observations exceeding the 4/n Cook's D threshold
cooksd_threshold <- 4 / nrow(naplan)
influential_idx <- which(final_aug$.cooksd > cooksd_threshold)
cat("Observations exceeding 4/n threshold:", length(influential_idx), "out of", nrow(naplan), "\n")

# Refit without them
naplan_trimmed <- naplan[-influential_idx, ]
mod_trimmed <- update(final_mod, data = naplan_trimmed)
```

```{webr ch-case-naplan-sensitivity-compare}
#| echo: true
#| autorun: true
# Compare key coefficients
comparison <- bind_rows(
  tidy(final_mod) |> mutate(model = "Full data"),
  tidy(mod_trimmed) |> mutate(model = "Trimmed")
) |>
  filter(term %in% c("genderMale", "school_typeIndependent",
                      "parent_educationPostgraduate", "parent_educationBachelor degree")) |>
  select(model, term, estimate, std.error) |>
  arrange(term, model)

comparison |> print(n = 20)
```

If coefficients are stable (only small changes in estimates and standard errors), our conclusions are robust. If they shift substantially, that would warrant a deeper investigation into what's driving those influential points.

::: Exercise
Refit the final model after removing observations where Cook's distance exceeds 4/n. Store the trimmed model as `mod_sensitivity`. Compare the coefficient for `genderMale` between the full and trimmed models — did it change meaningfully?

::: {.cell exercise='ex_cs_sensitivity' envir='Ex5' define='["mod_sensitivity"]'}
```{webr}
#| exercise: ex_cs_sensitivity
#| envir: Ex5
#| autorun: true
#| define:
#|   - mod_sensitivity
threshold <- 4 / nrow(naplan)
keep <- final_aug$.cooksd <= threshold
mod_sensitivity <- update(final_mod, data = naplan[keep, ])

# Compare gender coefficient
cat("Full model genderMale:", round(coef(final_mod)["genderMale"], 2), "\n")
cat("Trimmed model genderMale:", round(coef(mod_sensitivity)["genderMale"], 2), "\n")
```
:::

:::: {.solution exercise="ex_cs_sensitivity"}
::: {.cell exercise='ex_cs_sensitivity' solution='true'}
```{webr}
#| exercise: ex_cs_sensitivity
#| solution: true
threshold <- 4 / nrow(naplan)
keep <- final_aug$.cooksd <= threshold
mod_sensitivity <- update(final_mod, data = naplan[keep, ])

# Compare gender coefficient
cat("Full model genderMale:", round(coef(final_mod)["genderMale"], 2), "\n")
cat("Trimmed model genderMale:", round(coef(mod_sensitivity)["genderMale"], 2), "\n")
cat("Change:", round(coef(mod_sensitivity)["genderMale"] - coef(final_mod)["genderMale"], 2), "points\n")
```
:::
::::

::: {.cell exercise='ex_cs_sensitivity' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_cs_sensitivity
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  if (!exists("mod_sensitivity", envir = .envir_result)) fail("Create an object named `mod_sensitivity`.")
  m <- get("mod_sensitivity", envir = .envir_result)
  if (!inherits(m, "lm")) fail("`mod_sensitivity` should be an `lm` model.")
  if (nobs(m) >= nrow(naplan)) fail("The trimmed model should have fewer observations than the full model.")
  pass("Good — with n=3000, removing a handful of influential points barely changes the coefficients. The model is robust.")
})
```
:::
:::

## Reporting the results {#sec-reporting-the-results}

We now summarise the final model in a way that a reader can understand (see @sec-partial-regression-coefficients-betai and @sec-inference-about-the-response-y).

### Model summary

We fitted a multiple linear regression model to `r naplan_n` students. The overall model fit was:

- Adjusted $R^2 \approx$ `r round(final_glance$adj.r.squared, 3)`
- Residual standard error $\approx$ `r round(final_glance$sigma, 2)`
- Overall $F$-test: $F($`r final_fstat[2]`$, $`r final_fstat[3]`$) = `r round(final_fstat[1], 1)`, p `r format.pval(final_p_overall, digits = 2, eps = 1e-16)`

```{webr ch-case-naplan-final-fit}
#| echo: true
#| autorun: true
formula(final_mod)
glance(final_mod)
```

### Interpreting key coefficients

For categorical predictors, coefficients are interpreted relative to their reference levels (see @sec-mlr_cat). In this model the references are:

- `grade`: Year 3
- `parent_education`: Year 10 or below
- `school_type`: Government
- `gender`: Female

```{webr ch-case-naplan-final-coefs}
#| echo: true
#| autorun: true
final_terms |>
  filter(str_detect(term, "parent_education|school_type|gender")) |>
  print()
```

From the table:

- Higher parent education is associated with higher reading scores (e.g., Bachelor vs Year 10 or below: about `r round(coef(final_mod)["parent_educationBachelor degree"], 1)` points higher, holding other variables constant).
- Independent schools score higher than Government schools by about `r round(coef(final_mod)["school_typeIndependent"], 1)` points on average (Catholic vs Government is much smaller).
- Males score about `r round(coef(final_mod)["genderMale"], 1)` points lower than females, on average.

### Interpreting the grade × reading-time pattern

Because the model includes an interaction between grade and a quadratic reading-time term, it’s often easiest to interpret with predicted curves rather than trying to “read” raw polynomial coefficients (see @sec-continuous-by-categorical-interactions).

```{webr ch-case-naplan-effects-reading-time}
#| echo: true
#| autorun: true
profile_ref <- tibble(
  parent_education = factor("Year 12", levels = levels(naplan$parent_education)),
  school_type = factor("Government", levels = levels(naplan$school_type)),
  gender = factor("Female", levels = levels(naplan$gender))
)

grid <- expand_grid(
  grade = levels(naplan$grade),
  reading_time_home = seq(0, 120, by = 5)
) |>
  mutate(
    parent_education = profile_ref$parent_education,
    school_type = profile_ref$school_type,
    gender = profile_ref$gender
  )

grid$fit <- predict(final_mod, newdata = grid)

ggplot(grid, aes(x = reading_time_home, y = fit, colour = grade)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Reading time at home (minutes/week)",
    y = "Predicted NAPLAN reading score",
    colour = "Grade",
    caption = "Predictions shown for: Government school, Female, parent education = Year 12."
  )
```

### A concrete prediction (with a prediction interval)

Consider a Year 9 female student in a Government school, with parent education = Bachelor degree, who reports 30 minutes of reading at home per week. The fitted mean is:

```{webr ch-case-naplan-prediction}
#| echo: true
#| autorun: true
pred_profile <- tibble(
  grade = factor("Year 9", levels = levels(naplan$grade)),
  reading_time_home = 30,
  parent_education = factor("Bachelor degree", levels = levels(naplan$parent_education)),
  school_type = factor("Government", levels = levels(naplan$school_type)),
  gender = factor("Female", levels = levels(naplan$gender))
)

predict(final_mod, newdata = pred_profile, interval = "confidence")
predict(final_mod, newdata = pred_profile, interval = "prediction")
```

### Sensitivity check: what if we use the SES index instead?

If we swap `parent_education` for the numeric SES index, the SES index is strongly associated with score.

```{webr ch-case-naplan-ses-index-effect}
#| echo: true
#| autorun: true
tidy(mod_ses_index, conf.int = TRUE) |>
  filter(term == "ses_index") |>
  print()
```

This illustrates an important modelling lesson: how you **operationalise** a construct like “SES” can change both interpretation and model fit. In this dataset, parent education seems to capture SES-related variation particularly well.

::: Exercise
Create a prediction interval for a specific student profile.

Use a Year 9 student with:

- `reading_time_home = 30`
- `parent_education = "Bachelor degree"`
- `school_type = "Government"`
- `gender = "Female"`

Store the result of `predict(..., interval = "prediction")` in `pred_pi`.

::: {.cell exercise='ex_5.7' envir='Ex5' define='["pred_pi"]'}
```{webr}
#| exercise: ex_5.7
#| envir: Ex5
#| autorun: true
#| define:
#|   - pred_pi
my_profile <- tibble(
  grade = factor("Year 9", levels = levels(naplan$grade)),
  reading_time_home = 30,
  parent_education = factor("Bachelor degree", levels = levels(naplan$parent_education)),
  school_type = factor("Government", levels = levels(naplan$school_type)),
  gender = factor("Female", levels = levels(naplan$gender))
)

pred_pi <- predict(final_mod, newdata = my_profile, interval = "prediction")
pred_pi
```
:::

:::: {.solution exercise="ex_5.7"}
::: {.cell exercise='ex_5.7' solution='true'}
```{webr}
#| exercise: ex_5.7
#| solution: true
my_profile <- tibble(
  grade = factor("Year 9", levels = levels(naplan$grade)),
  reading_time_home = 30,
  parent_education = factor("Bachelor degree", levels = levels(naplan$parent_education)),
  school_type = factor("Government", levels = levels(naplan$school_type)),
  gender = factor("Female", levels = levels(naplan$gender))
)

pred_pi <- predict(final_mod, newdata = my_profile, interval = "prediction")
pred_pi
```
:::
::::

::: {.cell exercise='ex_5.7' envir='Ex5' check='true' class='wait'}
```{webr}
#| exercise: ex_5.7
#| envir: Ex5
#| check: true
#| class: wait
grade_this({
  if (!exists("pred_pi", envir = .envir_result)) fail("Create an object named `pred_pi`.")
  ans <- get("pred_pi", envir = .envir_result)
  if (!is.matrix(ans) || nrow(ans) != 1 || !all(c("fit", "lwr", "upr") %in% colnames(ans))) {
    fail("`pred_pi` should be a 1×3 matrix with columns `fit`, `lwr`, and `upr`.")
  }

  expected_profile <- tibble(
    grade = factor("Year 9", levels = levels(naplan$grade)),
    reading_time_home = 30,
    parent_education = factor("Bachelor degree", levels = levels(naplan$parent_education)),
    school_type = factor("Government", levels = levels(naplan$school_type)),
    gender = factor("Female", levels = levels(naplan$gender))
  )
  expected <- predict(final_mod, newdata = expected_profile, interval = "prediction")

  if (isTRUE(all.equal(ans, expected, tolerance = 1e-6))) {
    pass("Great — you computed the prediction interval correctly.")
  }
  fail("Not quite. Hint: check the profile values and that you used `interval = \"prediction\"`.")
})
```
:::
:::

## Conclusion {#sec-case-naplan-conclusion}

Using `r naplan_n` students across Years 3, 5, 7, and 9, we fitted a multiple linear regression model relating NAPLAN reading scores to reading time at home, grade, parent education, school type, and gender. The model explained about `r round(100 * final_glance$adj.r.squared, 1)`% of the variation in scores (adjusted $R^2 \approx$ `r round(final_glance$adj.r.squared, 3)`), with a residual standard error of about `r round(final_glance$sigma, 1)` points.

After adjusting for other variables, higher parent education was associated with higher reading scores (e.g., Bachelor vs Year 10 or below: roughly `r round(coef(final_mod)["parent_educationBachelor degree"], 1)` points higher). Independent schools scored higher than Government schools by about `r round(coef(final_mod)["school_typeIndependent"], 1)` points on average, and males scored about `r abs(round(coef(final_mod)["genderMale"], 1))` points lower than females.

The relationship between reading time at home and score was not purely linear and differed by grade (a grade × quadratic reading-time interaction). Predicted curves suggested that increases in reading time are associated with higher predicted scores, but the shape of that association varies across schooling levels.

**Limitations:** These are associations from observational data. Potential clustering within schools and omitted variables (e.g., prior achievement, classroom factors) may affect estimates. Always interpret results in context and re-check diagnostics, as in @sec-regression-pitfalls-and-diagnostics.
