---
format:
  live-html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
  edit: false

embed-resources: true

filters:
  - custom-numbered-blocks

custom-numbered-blocks:
  styles:
    math:
      appearance: quartothmlike
    text:
      appearance: textbox
    callout:
      appearance: foldbox
  classes:
    Assumption:
      colors: ["#ffd8d8ff", "#ff2212ff"]
      style: callout
      numbered: false
    Example:
      colors: ["#BBDEFB", "#2196F3"]
      style: foldbox
      collapse: false
    Exercise:
      colors: ["#C8E6C9", "#4CAF50"]
      boxstyle: foldbox.simple
      collapse: false
    Note:
      colors: ["#fffce3ff", "#FFEB3B"]
      style: text
      numbered: false
    Key-term:
      colors: ["#D1C4E9", "#673AB7"]
      style: text
    Key-point:
      colors: ["#FFF9C4", "#FFEB3B"]
      boxstyle: foldbox.simple
      collapse: false
      numbered: false

webr:
  render-df: gt-interactive
  packages: 
    - dplyr
    - ggplot2
    - tidyr
    - purrr
    - broom
    - gt
    - MASS
  cell-options:
    echo: false
    warning: false
    message: false
    eval: true

resources:
  - Data
---
{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ./_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr ch-pitfalls-diagnostics-setup}
#| include: false
library(tidyverse)
library(broom)
library(MASS)
theme_set(theme_bw())
```

```{r ch-pitfalls-diagnostics-setup-r}
#| include: false
library(tidyverse)
library(broom)
library(MASS)
theme_set(theme_bw())
```

# Pitfalls, Diagnostics, and Remedies {#sec-regression-pitfalls-and-diagnostics}

In @sec-model-building-and-selection we saw that complex regression models can go wrong due to [overfitting]{.glossary term="Overfitting"}. But even simple regression analyses can be misleading if the linear model we are fitting does not adequately capture the data-generating process.

::: Example
### [Anscombe's quartet: same regression, different data]{#sec-anscombes-quartet}
Regression summaries can mask radically different data patterns. Consider the following four datasets, known as *Anscombe's quartet*:

:::: {.panel-tabset}
### Scatter plots
```{r ch-pitfalls-diagnostics-anscombe-setup}
#| edit: false
anscombe_long <- anscombe |>
  pivot_longer(
    cols = everything(),
    names_to = c(".value", "set"),
    names_pattern = "([xy])(\\d)"
  ) |>
  mutate(set = paste0("Set ", set))
  
ggplot(anscombe_long, aes(x = x, y = y)) +
  geom_point(size = 2) +
  facet_wrap(~ set, ncol = 2) 
```

### dataset
```{webr ch-pitfalls-diagnostics-anscombe-scatter}
#| edit: false
anscombe
```
::::

These data are clearly very different, howerver each  variable has the same summary statistics: 
```{webr ch-pitfalls-diagnostics-anscombe-summary}
#| edit: false
anscombe_long |>
  group_by(set) |>
  summarise(
    n = n(),
    mean_x = mean(x),
    sd_x = sd(x),
    mean_y = mean(y),
    sd_y = sd(y),
    cor_xy = cor(x, y)
  ) |>
  mutate(
    x = sprintf("%.2f (%.2f)", mean_x, sd_x),
    y = sprintf("%.2f (%.2f)", mean_y, sd_y),
    Correlation = sprintf("%.3f", cor_xy)
  ) |>
  dplyr::select(set, n, x, y, Correlation) |>
  gt(rowname_col = "set")  |>
  tab_spanner(label= "Mean(SD)", columns = c(x, y)) |> 
  cols_align("center") |>
  tab_stubhead(md("**Dataset**")) |>
    tab_style(
    style = cell_text(weight = "bold"),
    locations = list(
      cells_column_spanners(),
      cells_stub()
    )
  )
```

Furthermore,  fitting a simple linear regression model to each dataset yields identical estimates:

:::: {.panel-tabset}
### fitted lines
```{webr ch-pitfalls-diagnostics-anscombe-fitted-lines}
#| edit: false
#| include: false
#| echo: false

ggplot(anscombe_long, aes(x = x, y = y)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se= FALSE, linewidth = 0.8) +
  facet_wrap(~ set, ncol = 2) 
```

### Regression summary
```{webr ch-pitfalls-diagnostics-anscombe-regression}
#| edit: false
list(
  "Set 1" = lm(y1 ~ x1, anscombe),
  "Set 2" = lm(y2 ~ x2, anscombe),
  "Set 3" = lm(y3 ~ x3, anscombe),
  "Set 4" = lm(y4 ~ x4, anscombe)
) |>
  map_df(~ tibble(
    Intercept = coef(.x)[1],
    Slope = coef(.x)[2],
    R_squared = summary(.x)$r.squared,
    Residual_SE = summary(.x)$sigma
  ), .id = "Dataset") |>
  gt() |>
  cols_label(
    Dataset = "Dataset",
    Intercept = "Intercept",
    Slope = "Slope",
    R_squared = "R²",
    Residual_SE = "Residual SE"
  ) |>
  fmt_number(
    columns = c(Intercept, Slope, Residual_SE),
    decimals = 2
  ) |>
  fmt_number(
    columns = R_squared,
    decimals = 3
  ) |>
  tab_style(
    style = cell_text(align = "center"),
    locations = cells_body(columns = c(Intercept, Slope, R_squared, Residual_SE))
  )
```

::::

The key lesson from Anscombe's quartet is that regression summaries can hide important aspects of the data. Therefore, a necessary step in **any** regression analysis is to check whether the data meet the assumptions of the linear model being fitted. 
:::

In this chapter we focus on the ways in which linear regression can go wrong, how to diagnose problems using residuals, and what to do when issues arise. 

::: Key-point
A good regression analysis is not just about fitting a model and reporting results; it also involves checking that the model assumptions are reasonably satisfied and addressing any issues that arise.
:::

- [ ] Continue

## Revisiting linear model assumptions {#sec-linear-model-and-its-assumptions}

In @sec-introduction-to-linear-models, we introduced the basic assumptions for constructing a simple linear model of the relationship between two variables. In @sec-multiple-linear-regression-mlr we extended these assumptions to the multiple predictor setting. As a reminder, a linear regression analysis assumes that the data were generated according to something of the following form:

::: Assumption
### The linear model {#assumption-linear-model}
The response $Y$ for observation $i$ is related to predictor variables $X_1, X_2, \ldots, X_p$ according to:
$$
Y_i = \beta_0 + \beta_1 X_{i,1} + \beta_2 X_{i,2} + \cdots + \beta_p X_{i,p} + \varepsilon_i, \quad i = 1, 2, \ldots, n
$$
where
$$
\varepsilon_i \stackrel{\text{iid}}{\sim} \text{Normal}(0, \sigma^2)
$$
for some unknown parameters $\beta_0, \beta_1, \ldots, \beta_p$ and $\sigma^2 > 0$.
:::

By assuming this form we can proceed to estimate the parameters $b_0, b_1, \ldots, b_p$ and $s^2$ using the method of least squares, and then use these estimates to perform inference and prediction. 

However, if it is not the case that the data were generated according to the proposed linear model (e.g. as in sets 2, 3, or 4 of Anscombe's quartet), then our estimates, inferences, and predictions may be misleading.

- [ ] Continue

### Common assumptions {#sec-common-assumptions-and-pitfalls}
It is often the case that some aspects of the proposed linear model may be violated while others are still reasonably satisfied. For example, data may show a linear trend but have nonconstant variance (heteroskedasticity). For this reason, we often break the general assumption above into more specific parts when performing model diagnostics. A common, equivalent way to present the assumptions of a linear model is via the following four assumptions:

::: Assumption
### Linearity
The relationship between the predictors and the response is linear (in the parameters) on average
$$
E[Y_i] = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip}
$$
:::

::: Assumption
### Homoscedasticity
The variance of the errors is constant across the range of predictors:
$$
\text{Var}(\epsilon_i) = \sigma^2
$$
:::
::: Assumption
### Normality
The errors are approximately normally distributed:
$$
\epsilon_i \sim \text{Normal}(0, \sigma^2)
$$
:::
::: Assumption
### Independence
The errors are independent: $\epsilon_i$ is independent of $\epsilon_j$ for $i \neq j$. 
:::

The standard workflow for a regression analysis involves checking whether these assumptions are reasonably satisfied. If problems are detected, it means we need to revise our model in some way (e.g., by adding polynomial terms, transforming variables, or using a different modelling approach).

- [ ] Continue

## Residuals as diagnostic tools {#sec-residuals-as-diagnostic-tools}

How can we check whether the assumptions of a linear regression model are met?
We might start by plotting the data and looking for obvious problems. This is how we identified issues in @sec-anscombes-quartet looking at Anscombe's quartet. However, a more systematic approach is to examine the *residuals* from the fitted model. 
Remember that the *residuals* are defined as the difference between the observed values and the fitted values from the model.

::: Key-term
### Residual {#sec-residual}
The residual for observation $i$ in a regression model is defined as
$$e_i = y_i - \hat{y}_i$$
where $e_i$ is the residual for observation $i$, $y_i$ is the observed value, and $\hat{y}_i$ is the predicted value from the model.
:::

We used residuals in @sec-leastsquares to estimate the parameters of the linear model, but they can also be used to diagnose model problems. Since residuals encode the information about what the model has *not* captured, plotting residuals can reveal systematic patterns where the model is failing to fit the data well.


### Standardised residuals, leverage, and Cook's distance {#sec-standardised-residuals-leverage-cook-s-distance}

Raw residuals $e_i$ are measured on the scale of the response, so their magnitude depends on the units of $Y$ (and on how variable $Y$ is). In multiple regression, residuals also do not all have the same variance: observations with unusual predictor values tend to have smaller residual variance simply because the fitted model is "pulled" towards them. To compare residuals on a common scale, we use *standardised residuals*.

::: Key-term
#### Standardised residual {#sec-standardised-residual}
The standardised residual for observation $i$ is
$$
r_i = \frac{e_i}{s\,\sqrt{1-h_{ii}}}
$$
where $s$ is the residual standard error from the fitted model and $h_{ii}$ is the [leverage] for observation $i$. Under the usual model assumptions, $r_i$ is approximately $\text{Normal}(0,1)$.
:::

Standardised residuals are especially useful when comparing residual sizes across observations, and they are the scale used in several common diagnostic plots. 

::: Key-term
#### Leverage (hat value) {#sec-leverage-hat-value}
Leverage measures how unusual an observation's predictor values are relative to the rest of the data. The leverage values $h_{ii}$ come from the *hat matrix* $H = X(X^\top X)^{-1}X^\top$, satisfy $0 \le h_{ii} \le 1$, and sum to $p + 1$ (including the intercept), so the average leverage is $(p + 1)/n$.

A common rule of thumb is to flag points with 
$$
h_{ii} > \frac{2(p + 1)}{n} \quad \text{(or } \frac{3(p + 1)}{n}\text{)}
$$
as "high leverage", where $p$ is the number of predictors (excluding the intercept) and $n$ is the sample size.
:::

High leverage does *not* automatically mean a problem: it becomes concerning when it coincides with a large standardised residual.

::: Key-term
#### Cook's distance {#sec-cooks-distance}
Cook's distance summarises how much the fitted model would change if observation $i$ were removed:
$$
D_i = \frac{e_i^2}{(p + 1)s^2}\,\frac{h_{ii}}{(1-h_{ii})^2}.
$$
Large values indicate potentially influential observations.
:::

::: Note
Rules of thumb (use with caution): observations with $|r_i| > 2$ (or $> 3$) may be unusual; Cook's distance values that stand out relative to the rest merit investigation.
:::

In practice, these diagnostics are easy to extract with `broom::augment()` (as `.std.resid`, `.hat`, and `.cooksd`).

::: Example
### Residuals in the Olympic cholesterol data {#sec-residuals-in-the-olympic-cholesterol-data}

The OLYMPIC dataset records fat intake (mg) and cholesterol (mg/L) for 20 athletes. 

```{webr ch-pitfalls-diagnostics-olympic-data}
#| edit: false

olympic <- read.delim("Data/OLYMPIC.txt")
```

::: {.panel-tabset}
## Scatter plot 
```{webr ch-pitfalls-diagnostics-olympic-data-scatter}
#| edit: false
ggplot(olympic, aes(x = FAT, y = CHOLESTEROL)) +
  geom_point(size = 2) +
  labs(x = "Fat intake (mg)", y = "Cholesterol (mg/L)")

```

## Data table
```{webr ch-pitfalls-diagnostics-olympic-data-table}
#| edit: false
olympic
```
:::

We can see that cholesterol tends to increase with fat intake, but the relationship does not appear to be perfectly linear. Let's fit a simple linear regression model predicting cholesterol from fat intake, and then examine the residuals to see if there are any obvious problems.

```{webr ch-pitfalls-diagnostics-olympic-setup}
#| edit: false
olympic_mod <- lm(CHOLESTEROL ~ FAT, data = olympic)
olympic$resid <- resid(olympic_mod)
olympic$fitted <- predict(olympic_mod)
```

:::: {.panel-tabset}
### Residuals vs fitted
```{webr ch-pitfalls-diagnostics-olympic-residuals-vs-fitted}
#| edit: false
ggplot(olympic, aes(x = fitted, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(size = 2) +
  labs(x = "Fitted value", y = "Residual")
```

### Residuals vs FAT
```{webr ch-pitfalls-diagnostics-olympic-residuals-vs-fat}
#| edit: false
ggplot(olympic, aes(x = FAT, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(size = 2) +
  labs(y = "Residual")
```
::::

We can see that the positive relationship between fat intake and cholesterol observed in the scatter plot has been captured by the model (the diagonal trend is gone in the residual plots, and the coefficient for FAT in the model summary is positive and significant). However, there is still a clear curved pattern in the residuals, indicating that not all the structure in the data has been captured.

The curved pattern in the data is not able to be captured by a first order linear model. This suggests that we need to revise our model, perhaps by adding a polynomial term or transforming one of the variables.
:::

### Extracting diagnostics with `broom::augment()` {#sec-extracting-residuals-and-broom-augment}

In @sec-residuals-as-diagnostic-tools we defined residuals; we can extract them using `resid()`. A convenient alternative is `broom::augment()`, which returns a data frame containing the original data plus fitted values, residuals, and several diagnostic quantities.

::: Example
### Olympic residuals with `broom::augment()` {#sec-olympic-residuals-with-broom-augment}

```{webr ch-pitfalls-diagnostics-olympic-augment-preview}
#| edit: false
olympic_mod |>
  augment() |> 
  select(FAT, CHOLESTEROL, .fitted, .resid, .std.resid, .hat, .cooksd)
```

The columns `.std.resid`, `.hat`, and `.cooksd` correspond to standardised residuals, leverage, and Cook's distance. One simple way to spot potentially influential points is to sort by Cook's distance:

```{webr ch-pitfalls-diagnostics-olympic-influence-table}
#| edit: false
olympic_mod |>
  augment() |>
  mutate(.index = row_number()) |>
  select(.index, FAT, CHOLESTEROL, .std.resid, .hat, .cooksd) |>
  arrange(desc(.cooksd)) |>
  slice(1:6)
```
:::

We will use the 'augmented' data frame format extensively in the remainder of this chapter.

- [ ] Continue

## Residual diagnostics by assumption {#sec-common-diagnostic-plots}

The goal of residual diagnostics is to check whether the key assumptions behind our linear regression model are reasonable.

We'll illustrate using a multiple linear regression model predicting miles per gallon (`mpg`) from weight (`wt`), horsepower (`hp`), and transmission type (`am`) in the `mtcars` dataset.

```{r ch-pitfalls-diagnostics-diagnostic-model-setup}
#| edit: false
diag_mod <- lm(mpg ~ wt + hp + am, data = mtcars)
```
We will use `broom::augment()` to extract the residuals and other diagnostics.
```{r ch-pitfalls-diagnostics-diagnostic-model-augment}
#| edit: false
diag_aug <- augment(diag_mod)
```

And proceed by checking each assumption from @sec-common-assumptions-and-pitfalls in turn.

### Linearity
Again, we assume that the mean of the response is a linear function of the predictors.

::: Assumption
The conditional mean of the response is linear in the predictors:
$$
E[Y_i] = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \cdots + \beta_p X_{ip}.
$$
:::

Because we make this assumption, our least squares fit finds the best straight line (in the least squares sense). If the data fit this assumption, then the residuals plotted against fitted values should look like random noise around 0. Systematic curvature suggests the mean structure is misspecified (e.g., a missing nonlinear term or interaction).

```{r ch-pitfalls-diagnostics-residuals-vs-fitted-ggplot}
#| edit: false
ggplot(diag_aug, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(size = 2) +
  labs(x = "Fitted value", y = "Residual")
```

### Constant variance
The vertical spread of residuals should be roughly constant across the range of fitted values. A "funnel" pattern suggests heteroskedasticity. The scale-location plot makes changes in spread easier to see.

```{r ch-pitfalls-diagnostics-scale-location-ggplot}
#| edit: false
ggplot(diag_aug, aes(x = .fitted, y = sqrt(abs(.std.resid)))) +
  geom_point(size = 2) +
  geom_smooth(se = FALSE, linewidth = 0.8, color = "gray40") +
  labs(x = "Fitted value", y = expression(sqrt("|Standardised residual|")))
```

### Normality
A Normal Q-Q plot compares the standardised residuals to what we'd expect under normal errors. Strong departures from the line suggest non-normality (often due to outliers or skewness).

```{webr ch-pitfalls-diagnostics-normal-qq-ggplot}
#| edit: false
ggplot(diag_aug, aes(sample = .std.resid)) +
  stat_qq(size = 2) +
  stat_qq_line(color = "gray40") +
  labs(x = "Theoretical quantiles", y = "Standardised residuals")
```

### Unusual and influential observations
Influence combines (i) leverage (unusual predictor values) and (ii) residual size. Points with high leverage and large residuals can have an outsized effect on the fitted model.

```{webr ch-pitfalls-diagnostics-residuals-vs-leverage-ggplot}
#| edit: false
ggplot(diag_aug, aes(x = .hat, y = .std.resid, size = .cooksd)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(alpha = 0.8) +
  scale_size_continuous(range = c(1, 10)) +
  labs(x = "Leverage (hat value)", y = "Standardised residual", size = "Cook's D")
```

::: Key-point
If residuals show a clear *pattern*, that pattern is telling you what your model is missing.
:::

### Shortcut: `plot()` for `lm` objects
Once you know what to look for, base R will produce the same set of standard diagnostic plots automatically (and you can request a specific panel with `which = 1`, `2`, `3`, or `5`).

```{webr ch-pitfalls-diagnostics-plot-lm}
#| edit: false
plot(diag_mod, ask = FALSE)
```

- [ ] Continue

## Diagnosing issues and choosing remedies {#sec-diagnosing-issues-and-choosing-remedies}
We can use the diagnostic plots above to identify common problems with linear regression models. Below we outline some typical issues, how to recognise them in residual plots, and common strategies for addressing them.

### Nonlinearity (missing curvature) {#sec-diagnosing-nonlinearity}

**Symptoms**

- A curved trend in residuals vs fitted (or residuals vs a predictor).

**Common remedies**

- Add polynomial terms (see @sec-non_linearity), transformations, or interactions (see @sec-interaction-models-with-continuous-predictors).
- Consider omitted predictors that explain structure in the residuals.

::: Example
#### A residual curve suggests missing nonlinearity {#sec-nonlinearity-example}

```{webr ch-pitfalls-diagnostics-nonlinearity-sim}
#| edit: false
set.seed(42)
n <- 80
x <- runif(n, -3, 3)
y <- 2 + 1.2 * x + 0.8 * x^2 + rnorm(n, 0, 2)
nl_dat <- tibble(x, y)

m_lin <- lm(y ~ x, data = nl_dat)
m_quad <- lm(y ~ x + I(x^2), data = nl_dat)

aug_lin <- augment(m_lin)
aug_quad <- augment(m_quad)
```

:::: {.panel-tabset}
### Linear model residuals
```{webr ch-pitfalls-diagnostics-nonlinearity-lin-resid}
#| edit: false
ggplot(aug_lin, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(size = 2) +
  labs(title = "Linear fit", x = "Fitted value", y = "Residual")
```

### Quadratic model residuals
```{webr ch-pitfalls-diagnostics-nonlinearity-quad-resid}
#| edit: false
ggplot(aug_quad, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(size = 2) +
  labs(title = "Quadratic fit", x = "Fitted value", y = "Residual")
```
::::
:::

### Nonconstant variance ([heteroskedasticity]{.glossary}) {#sec-diagnosing-heteroskedasticity}

**Symptoms**

- A “funnel” shape in residuals vs fitted (spread increases or decreases with the fitted value).
- A trend in the scale–location plot.

**Common remedies**

- Transform the response (e.g., log or square-root) to stabilise variance.
- Use weighted least squares if you can model how variance changes.
- Use robust standard errors when inference is the main goal.

::: Example
#### A funnel shape suggests heteroskedasticity {#sec-heteroskedasticity-example}

```{webr ch-pitfalls-diagnostics-hetero-sim}
#| edit: false
set.seed(7)
n <- 120
x <- runif(n, 0, 10)
y <- 20 + 3 * x + rnorm(n, 0, sd = 0.7 * x)
het_dat <- tibble(x, y)

m_raw <- lm(y ~ x, data = het_dat)
m_log <- lm(log(y) ~ x, data = het_dat)

aug_raw <- augment(m_raw)
aug_log <- augment(m_log)
```

:::: {.panel-tabset}
### Residuals (raw scale)
```{webr ch-pitfalls-diagnostics-hetero-raw}
#| edit: false
ggplot(aug_raw, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(size = 2) +
  labs(title = "Funnel-shaped residuals", x = "Fitted value", y = "Residual")
```

### Residuals (log-transformed response)
```{webr ch-pitfalls-diagnostics-hetero-log}
#| edit: false
ggplot(aug_log, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(size = 2) +
  labs(title = "After log transform", x = "Fitted value", y = "Residual")
```
::::
:::

### Normality (and when it matters) {#sec-assessing-normality-of-residuals}

Normality of errors matters most for small-sample $t$-tests and confidence intervals. In large samples, linear regression often remains useful even when residuals are not perfectly normal, but it is still worth checking for strong departures caused by outliers or skewness.

**Symptoms**

- Strong curvature or heavy tails in the Q–Q plot.

**Common remedies**

- Check for outliers/data errors first.
- Consider a transformation of the response.
- Consider robust or resampling-based inference.

### Independence (autocorrelation) {#sec-assessing-independence}

The [independence]{.glossary} assumption is most commonly violated when observations are collected over time (or space). If residuals are correlated, standard errors can be too small, and $p$-values can look “better” than they should.

**Symptoms**

- Residuals show runs/clustering when plotted against time/order.
- Residuals have clear autocorrelation (e.g., via an ACF plot).

**Common remedies**

- Add time/seasonality terms if they explain structure.
- Use models designed for correlated errors (time series, mixed models, GLS).

::: Example
#### Residuals over time can reveal dependence {#sec-independence-example}

```{webr ch-pitfalls-diagnostics-independence-sim}
#| edit: false
set.seed(99)
n <- 90
time <- seq_len(n)
x <- rnorm(n)

# AR(1) errors
phi <- 0.7
e <- numeric(n)
e[1] <- rnorm(1)
for (i in 2:n) e[i] <- phi * e[i - 1] + rnorm(1)

y <- 1 + 0.6 * x + e
dep_dat <- tibble(time, x, y)

dep_mod <- lm(y ~ x, data = dep_dat)
dep_res <- resid(dep_mod)

acf_obj <- acf(dep_res, plot = FALSE)
acf_df <- tibble(lag = as.numeric(acf_obj$lag), acf = as.numeric(acf_obj$acf)) |>
  filter(lag > 0)
```

:::: {.panel-tabset}
### Residuals vs time
```{webr ch-pitfalls-diagnostics-independence-resid-time}
#| edit: false
tibble(time = time, resid = dep_res) |>
  ggplot(aes(x = time, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_line(linewidth = 0.6) +
  labs(x = "Time (order)", y = "Residual")
```

### ACF of residuals
```{webr ch-pitfalls-diagnostics-independence-acf}
#| edit: false
ggplot(acf_df, aes(x = lag, y = acf)) +
  geom_hline(yintercept = 0, color = "gray40") +
  geom_segment(aes(xend = lag, yend = 0), linewidth = 0.8) +
  labs(x = "Lag", y = "Autocorrelation (ACF)")
```
::::
:::

### Extrapolation and high-leverage prediction {#sec-extrapolation-and-high-leverage}

[Extrapolation]{.glossary} means predicting for predictor values outside the range you observed. It is risky because the *relationship itself* may change outside the observed region, and because uncertainty grows rapidly when you move away from the data cloud (often showing up as high leverage).

::: Key-point
Before you trust a prediction, check whether it is an extrapolation. A simple first step is to compare each new predictor value to the observed range.
:::

```{webr ch-pitfalls-diagnostics-extrapolation-range}
#| edit: false
tibble(
  predictor = c("wt", "hp"),
  min = c(min(mtcars$wt), min(mtcars$hp)),
  max = c(max(mtcars$wt), max(mtcars$hp))
)
```

### Transformations including Box–Cox {#sec-transformations-including-box-cox}

Transformations can help when residual plots show curvature or nonconstant variance. A systematic way to explore power transformations of the response is the [Box–Cox transformation]{.glossary term="Box-Cox transformation"}.

::: Note
Box–Cox requires the response to be strictly positive.
:::

```{webr ch-pitfalls-diagnostics-boxcox}
#| edit: false
bc <- MASS::boxcox(diag_mod, plotit = FALSE)
bc_df <- tibble(lambda = bc$x, logLik = bc$y)
lambda_hat <- bc_df$lambda[which.max(bc_df$logLik)]

ggplot(bc_df, aes(x = lambda, y = logLik)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = lambda_hat, linetype = "dashed", color = "gray40") +
  labs(title = "Box–Cox profile log-likelihood",
       subtitle = paste0("Best lambda ≈ ", round(lambda_hat, 2)),
       x = expression(lambda),
       y = "Profile log-likelihood")
```

### Multicollinearity {#sec-multicollinearity-diagnostics}

[Multicollinearity]{.glossary} occurs when predictors are strongly correlated. It does *not* necessarily make predictions bad, but it can make coefficient estimates unstable and inflate standard errors—so interpretation and “which predictor matters?” questions become fragile.

**Symptoms**

- Large standard errors and coefficient sign changes when you add/remove a correlated predictor.
- Large [variance inflation factors]{.glossary term="Variance inflation factor"} (VIFs).

**Common remedies**

- Drop or combine redundant predictors (guided by scientific context).
- Collect more data (especially with more diverse predictor combinations).
- Use dimension reduction or penalised regression when prediction is the goal.

```{webr ch-pitfalls-diagnostics-vif-helper}
#| edit: false
vif_from_lm <- function(model) {
  x <- model.matrix(model)
  x <- x[, colnames(x) != "(Intercept)", drop = FALSE]
  if (ncol(x) == 0) return(tibble(term = character(), vif = numeric()))
  if (ncol(x) == 1) return(tibble(term = colnames(x), vif = 1))

  vifs <- map_dbl(seq_len(ncol(x)), function(j) {
    r2 <- summary(lm(x[, j] ~ x[, -j]))$r.squared
    1 / (1 - r2)
  })

  tibble(term = colnames(x), vif = vifs) |>
    arrange(desc(vif))
}
```

```{webr ch-pitfalls-diagnostics-vif-example}
#| edit: false
col_mod <- lm(mpg ~ wt + disp + hp, data = mtcars)
vif_from_lm(col_mod)
```

### Blind model selection {#sec-blind-model-selection}

Automated selection procedures (stepwise, “try everything”, repeatedly adding/removing predictors until something is significant) can be useful for exploration, but they are easy to misuse.

**Pitfalls**

- $p$-values and confidence intervals become hard to interpret after extensive searching.
- Selected models can overfit and can be unstable to small data changes.

**Remedies**

- Use scientific context to limit the candidate set.
- Use out-of-sample validation (see @sec-why-not-increase-model-complexity) and always re-check residual diagnostics on the final model.

- [ ] Continue

## Exercises {#sec-pitfalls-exercises}

The goal of these exercises is to practise diagnosing patterns and choosing a sensible next step.

### Exercise 1: Identify heteroskedasticity {#sec-pitfalls-exercise-1}

Run the simulation below and look at the residual plot. Then answer: what issue is the plot mainly suggesting?

```{webr ch-pitfalls-diagnostics-ex1-setup}
#| echo: false
#| edit: false
set.seed(123)
n <- 150
x <- runif(n, 0, 10)
y <- 10 + 2 * x + rnorm(n, 0, sd = 0.4 * x)
ex1_dat <- tibble(x, y)
ex1_mod <- lm(y ~ x, data = ex1_dat)
ex1_aug <- augment(ex1_mod)

ggplot(ex1_aug, aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_point(size = 2) +
  labs(x = "Fitted value", y = "Residual")
```

::: Exercise
Write your answer as a short phrase (e.g., `"heteroskedasticity"` or `"nonconstant variance"`).

::: {.cell exercise='ex_4.1' envir='Ex4' define='["issue_4_1"]'}
```{webr}
#| exercise: ex_4.1
#| envir: Ex4
#| define:
#|   - issue_4_1
issue_4_1 <- "____"
```
:::

:::: {.solution exercise="ex_4.1"}
::: {.cell exercise='ex_4.1' solution='true'}
```{webr}
#| exercise: ex_4.1
#| solution: true
issue_4_1 <- "heteroskedasticity"
```
:::
::::

::: {.cell exercise='ex_4.1' envir='Ex4' check='true' class='wait'}
```{webr}
#| exercise: ex_4.1
#| envir: Ex4
#| check: true
#| class: wait
grade_this({
  ans <- tolower(trimws(get("issue_4_1", envir = .envir_result)))
  ok <- ans %in% c("heteroskedasticity", "nonconstant variance", "heteroscedasticity")
  if (ok) pass("Yes — the spread increases with the fitted value.")
  fail("Hint: look for a funnel shape (changing spread).")
})
```
:::
:::

- [ ] Continue

### Exercise 2: Use VIF to spot multicollinearity {#sec-pitfalls-exercise-2}

Fit the model and compute VIFs. Which term has the largest VIF?

```{webr ch-pitfalls-diagnostics-ex2-setup}
#| echo: false
#| edit: false
ex2_mod <- lm(mpg ~ wt + disp + hp, data = mtcars)
vif_from_lm(ex2_mod)
```

::: Exercise
Set `worst_term` equal to the term with the largest VIF.

::: {.cell exercise='ex_4.2' envir='Ex4' define='["worst_term"]'}
```{webr}
#| exercise: ex_4.2
#| envir: Ex4
#| define:
#|   - worst_term
worst_term <- "____"
```
:::

:::: {.solution exercise="ex_4.2"}
::: {.cell exercise='ex_4.2' solution='true'}
```{webr}
#| exercise: ex_4.2
#| solution: true
worst_term <- vif_from_lm(ex2_mod)$term[1]
```
:::
::::

::: {.cell exercise='ex_4.2' envir='Ex4' check='true' class='wait'}
```{webr}
#| exercise: ex_4.2
#| envir: Ex4
#| check: true
#| class: wait
grade_this({
  true_term <- vif_from_lm(ex2_mod)$term[1]
  ans <- trimws(get("worst_term", envir = .envir_result))
  if (identical(ans, true_term)) pass("Correct.")
  fail("Hint: your answer should match the first row of the VIF table (largest VIF).")
})
```
:::
:::

- [ ] Continue

### Exercise 3: Interpret a Box–Cox suggestion {#sec-pitfalls-exercise-3}

In the Box–Cox plot earlier, we computed `lambda_hat`. Based on its value, which option is the best *rough* interpretation?

::: Exercise
Choose one: `"no transform"`, `"log transform"`, or `"square-root transform"`.

::: {.cell exercise='ex_4.3' envir='Ex4' define='["boxcox_choice"]'}
```{webr}
#| exercise: ex_4.3
#| envir: Ex4
#| define:
#|   - boxcox_choice
boxcox_choice <- "____"
```
:::

:::: {.solution exercise="ex_4.3"}
::: {.cell exercise='ex_4.3' solution='true'}
```{webr}
#| exercise: ex_4.3
#| solution: true
boxcox_choice <- if (abs(lambda_hat) < 0.25) "log transform" else if (abs(lambda_hat - 0.5) < 0.25) "square-root transform" else "no transform"
```
:::
::::

::: {.cell exercise='ex_4.3' envir='Ex4' check='true' class='wait'}
```{webr}
#| exercise: ex_4.3
#| envir: Ex4
#| check: true
#| class: wait
grade_this({
  ans <- tolower(trimws(get("boxcox_choice", envir = .envir_result)))
  expected <- if (abs(lambda_hat) < 0.25) "log transform" else if (abs(lambda_hat - 0.5) < 0.25) "square-root transform" else "no transform"
  if (identical(ans, expected)) pass("Good — you matched the rough interpretation of lambda.")
  fail("Hint: lambda ≈ 0 suggests log; lambda ≈ 0.5 suggests sqrt; lambda ≈ 1 suggests no transform.")
})
```
:::
:::

## Handling [outliers]{.glossary term="Outlier"} and influential observations {#sec-handling-outliers-and-influential-observations}

When diagnostics flag influential observations, slow down and investigate rather than deleting points automatically.

- Check data quality first (entry errors, unusual units, miscoding).
- If influential points are real, fit with and without them to assess robustness; report how conclusions change.
- Prefer model improvements (functional form, additional predictors, variance stabilisation) over deletion.
