---
title: "Interaction and Nonlinearity"
format:
  live-html:
    toc: true
---

## Interaction models with quantitative predictors

- Allow the effect of one predictor to depend on another
  ([interaction](#gloss-interaction)):
  $$E[Y] = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1X_2.$$
- $\beta_3$ shifts the slope of $X_1$ per-unit change in $X_2$ (and
  vice versa).

```{r}
int_mod <- lm(mpg ~ wt * hp, data = mtcars)
coef(int_mod)[c("wt", "hp", "wt:hp")]
```

## Graphical interpretation of interaction effects

- Plot fitted lines across a grid to see slope changes.

```{r}
library(ggplot2)
grid <- expand.grid(wt = seq(2, 4, 0.5), hp = c(90, 150))
grid$fit <- predict(int_mod, grid)
ggplot(grid, aes(wt, fit, colour = factor(hp))) +
  geom_line() + labs(colour = "HP level", y = "Fitted mpg")
```

## Polynomial models: quadratic and cubic

- Capture curvature by adding powers: $$E[Y] = \beta_0 + \beta_1 X +
  \beta_2 X^2 \;(+\; \beta_3 X^3).$$
- Use `poly()` or explicit powers; center $X$ to reduce collinearity.

```{r}
poly_mod <- lm(mpg ~ wt + I(wt^2), data = mtcars)
```

## When and how to model curvature

- Use scatterplots and residual-vs-fitted plots to spot nonlinearity.
- Prefer low-order polynomials for interpretability; consider splines
  for flexible shapes if allowed.

## Extrapolation risks and overfitting concerns

- Polynomial terms can explode outside the data range—avoid predicting
  far beyond observed $X$.
- Guard against overfitting with cross-validation or an independent
  validation set when sample size permits.
