---
title: "Qualitative Predictors"
format:
  live-html:
    toc: true
---

## Dummy-variable coding

- Represent $k$-level categorical predictors with $k-1$ indicator
  variables ([dummy variables](#gloss-dummy)); the omitted level is the
  baseline.

```{r}
cat_mod <- lm(mpg ~ factor(cyl), data = mtcars)
model.matrix(cat_mod)[1:5, ]
```

## Baseline category interpretation

- Each indicator coefficient compares its level to the baseline.
- Re-level with `relevel()` for more meaningful comparisons.

```{r}
mtcars$cyl <- relevel(factor(mtcars$cyl), ref = "6")
relevel_mod <- lm(mpg ~ cyl, data = mtcars)
coef(relevel_mod)
```

## Regression with multi-level factors

- Fit models with multiple factors and quantitative predictors; ensure
  design matrix is full rank (no redundant indicators).

```{r}
mix_mod <- lm(mpg ~ wt + factor(carb), data = mtcars)
```

## Mixing categorical and continuous predictors

- Combine factors and continuous terms; interaction terms allow
  different slopes by group (e.g., `wt * cyl`).

```{r}
group_slope <- lm(mpg ~ wt * factor(gear), data = mtcars)
```

## Connection to ANOVA-style hypotheses

- ANOVA table for a factor tests whether any level differs from the
  baseline (joint $H_0$ on all indicators).
- In R, compare models with `anova()` or read the factor-level F-test in
  `summary()` output when using treatment coding.
