---
format:
  live-html:
    toc: true
---
# Regression Case Study: NAPLAN Reading Scores

## Defining the research question

- Investigate how student and school-level predictors relate to NAPLAN
  Reading scores. Clarify outcome, units, and any grouping structure.

## Importing data

```{r}
#| message: false
# Replace path with the appropriate CSV or RDS from the Resources/naplan reading folder
# naplan <- read.csv("Resources/naplan reading/naplan_reading.csv")
```

## Exploratory plot

- Start with scatterplots and boxplots for key predictors (e.g., study
  time, SES, gender) against Reading scores.

```{r}
# Example placeholder using mtcars structure; swap to naplan data
library(ggplot2)
ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  labs(x = "Predictor", y = "Reading score")
```

## Fitting the least-squares model

- Begin with a first-order additive model; consider interactions or
  polynomials if exploratory plots suggest curvature.

```{r}
# model <- lm(Reading ~ predictor1 + predictor2, data = naplan)
# summary(model)
```

## Model refinement and interpretation

- Check residual diagnostics (see Module 6). Address nonlinearity,
  heteroskedasticity, or influential points.
- Interpret coefficients, including categorical contrasts and any
  interaction terms.

```{r}
# plot(model, which = 1:2)
# coef(summary(model))
```

## Reporting the results

- Present fitted effects with confidence intervals, and provide a
  prediction interval for a meaningful scenario (e.g., a student profile
  of interest).
- Summarise key findings in plain language and note any limitations
  (e.g., omitted variables, sample size).
